---
title: "Lab: One categorical predictor"
execute:
  eval: false
---

In this lab, you will practice fitting regression models with one categorical predictor. Analyze the STePS-study data and document your findings in a Quarto notebook.

# Scenario

Your supervisor hasn't responded to your email about reading the updated version of your ISP. This is expected, after all, since they are a highly cited researcher with many ongoing projects. While working on the main project of your PhD studies, your supervisor suddenly appears at your desk:

"Hey, at the conference last week I started a collaboration with a colleague from Prestige University. We both thought it would be exciting to evaluate the relationship between education and LSAS-scores in the STePS-study. Could you help me with that?"

Use the tools you have learned about in the previous chapters to analyze the data. Write down your findings and data analysis decisions in a Quarto notebook.

# Setup

## Download data

Make sure you have downloaded the STePS-study data, and added it to the `data` folder inside your project. You can find the dataset on canvas.

## Load packages

```{r}
library(tidyverse) # Data wrangling and plotting
library(broom) # Converting model objects into tidy tables
library(quantreg) # Quantile regression (rq())
library(marginaleffects) # Predictions and uncertainty estimates
library(here) # File path management
```

Load the data into your project.

```{r}
df_rawdata <- read_csv("data/steps_clean.csv")
```

## Check data structure

Do a quick check of the data structure to verify that the CSV-file was loaded correctly.

```{r}
glimpse(df_rawdata)
```

# Data cleaning

Are there any data cleaning steps you need to take? Check for missing or weird values, odd responses succh as extreme outliers or implausible values.

Other types of data cleaning might involve creating new variables with simpler or new categories, or new ways to code existing variables.

# Analysis

Up to you. Look at the resources in the course chapters for help. Try to fit at least one model, extract the results, and create a plot showing your findings.
