---
title: "Logistic regression with two binary predictors"
format:
  html:
    df-print: kable
toc: true
toc-depth: 3
number-sections: true
---

# Two predictors, one clinical question

The previous logistic-regression chapters considered one predictor at a time. Clinical questions often ask whether a treatment contrast looks different in a baseline subgroup. Here we model the probability of a **low post-treatment LSAS score** using:

- treatment group (`trt`); and
- a binary baseline PHQ-9 category (`phq_cat`).

Our teaching question is:

> Within each PHQ-9 category, how does the estimated probability of a low post-treatment LSAS score differ between therapist-guided treatment and waitlist?

We will first fit an additive logistic model and then a model with `trt * phq_cat`. The probability difference, with its 95% confidence interval, is the clinical contrast. Log-odds and odds ratios help us read model output and research papers, but neither is a percentage-point difference.

# Packages and data

```{r}
#| label: setup
#| message: false
#| warning: false
library(tidyverse)
library(here)
library(broom)
library(marginaleffects)

options(
  marginaleffects_print_omit = c("s.value"),
  marginaleffects_print_style = ifelse(interactive(), "summary", "tinytable")
)
```

```{r}
#| label: load-and-derive-data
#| message: false
#| warning: false
df_binary <- read_csv2(here("data", "steps_clean.csv")) |>
  filter(!is.na(lsas_post)) |>
  mutate(
    # The second outcome level is the event modelled by binomial glm().
    lsas_post_bin = factor(
      if_else(lsas_post < 50, "Low", "High"),
      levels = c("High", "Low")
    ),
    trt = factor(
      trt,
      levels = c("waitlist", "therapist-guided", "self-guided")
    ),
    phq_cat = factor(
      case_when(
        phq9_screen < 10 ~ "< 10",
        phq9_screen >= 10 ~ ">= 10"
      ),
      levels = c("< 10", ">= 10")
    )
  )

df_binary |>
  count(phq_cat, lsas_post_bin, .drop = FALSE)
```

`phq_cat` explicitly divides baseline PHQ-9 scores into `< 10` and `>= 10`. It is a convenient binary moderator for this lesson, not a claim that these two groups are biologically distinct or that a subgroup result is clinically established. The outcome is also the Course 1 teaching outcome: `Low` when post-treatment LSAS is `< 50` and `High` when it is `>= 50`.

Waitlist is the reference treatment group, `< 10` is the reference PHQ-9 category, and `Low` is the modelled outcome event.

# Additive logistic model

An additive model has treatment and PHQ-9-category terms but no product term:

$$
\operatorname{logit}\left[P(Y_i = \mathrm{Low} \mid trt_i, phq\_cat_i)\right] =
\beta_0 + \beta_{trt} + \beta_{phq}.
$$

On the **log-odds scale**, this model says that the treatment odds ratio is the same in both PHQ-9 categories. That does *not* guarantee identical probability differences: changing a probability by a fixed odds ratio can produce different percentage-point changes at different baseline probabilities.

```{r}
#| label: fit-additive-logit
#| message: false
#| warning: false
mod_additive <- stats::glm(
  lsas_post_bin ~ trt + phq_cat,
  family = binomial,
  data = df_binary
)

additive_probabilities <- avg_predictions(
  mod_additive,
  by = c("trt", "phq_cat"),
  type = "response"
)

additive_treatment_effects <- avg_comparisons(
  mod_additive,
  variables = list(trt = c("waitlist", "therapist-guided")),
  by = "phq_cat",
  type = "response"
)

additive_treatment_low_phq <- additive_treatment_effects |>
  filter(phq_cat == "< 10")
additive_treatment_high_phq <- additive_treatment_effects |>
  filter(phq_cat == ">= 10")

additive_odds_ratio <- tidy(
  mod_additive,
  conf.int = TRUE,
  exponentiate = TRUE
) |>
  filter(term == "trttherapist-guided")

summary(mod_additive)
additive_probabilities
additive_treatment_effects
```

::: {.callout-tip collapse="true"}
## Model interpretation

Compared with waitlist, therapist-guided treatment has an estimated probability of a low post-treatment LSAS score that is **`r if (additive_treatment_low_phq$estimate[1] >= 0) "higher" else "lower"` by `r sprintf("%.1f", 100 * abs(additive_treatment_low_phq$estimate[1]))` percentage points** among participants with baseline PHQ-9 `< 10` (95% CI **`r sprintf("%.1f", 100 * additive_treatment_low_phq$conf.low[1])` to `r sprintf("%.1f", 100 * additive_treatment_low_phq$conf.high[1])` percentage points**). Among participants with baseline PHQ-9 `>= 10`, it is **`r if (additive_treatment_high_phq$estimate[1] >= 0) "higher" else "lower"` by `r sprintf("%.1f", 100 * abs(additive_treatment_high_phq$estimate[1]))` percentage points** (95% CI **`r sprintf("%.1f", 100 * additive_treatment_high_phq$conf.low[1])` to `r sprintf("%.1f", 100 * additive_treatment_high_phq$conf.high[1])` percentage points**).

Because this is an additive *logistic* model, its therapist-guided-versus-waitlist odds ratio is **`r sprintf("%.2f", additive_odds_ratio$estimate[1])`** (95% CI **`r sprintf("%.2f", additive_odds_ratio$conf.low[1])` to `r sprintf("%.2f", additive_odds_ratio$conf.high[1])`**) in both PHQ-9 categories, adjusted for PHQ-9 category. This is a ratio of odds, not a percentage-point probability change; the conditional probability differences above are the primary results.
:::

```{r}
#| label: plot-additive-probabilities
#| fig-cap: "Additive-model estimated probabilities of a low post-treatment LSAS score. Error bars are 95% confidence intervals."
plot_predictions(
  mod_additive,
  by = c("trt", "phq_cat"),
  type = "response"
) +
  labs(
    title = "Additive logistic model",
    subtitle = "Estimated probabilities by treatment and baseline PHQ-9 category",
    x = "Treatment group",
    y = "Estimated probability of low post-treatment LSAS (95% CI)",
    colour = "Baseline PHQ-9"
  ) +
  theme_minimal()
```

# Interaction model

The interaction model uses the requested product term, `trt * phq_cat`:

$$
\operatorname{logit}\left[P(Y_i = \mathrm{Low} \mid trt_i, phq\_cat_i)\right] =
\beta_0 + \beta_{trt} + \beta_{phq} + \beta_{trt \times phq}.
$$

The interaction terms let treatment associations with the outcome differ by baseline PHQ-9 category. The treatment main-effect coefficients are now interpreted at the reference category, PHQ-9 `< 10`; they are not an overall treatment effect.

```{r}
#| label: fit-interaction-logit
#| message: false
#| warning: false
mod_interaction <- stats::glm(
  lsas_post_bin ~ trt * phq_cat,
  family = binomial,
  data = df_binary
)

interaction_probabilities <- avg_predictions(
  mod_interaction,
  by = c("trt", "phq_cat"),
  type = "response"
)

interaction_treatment_effects <- avg_comparisons(
  mod_interaction,
  variables = list(trt = c("waitlist", "therapist-guided")),
  by = "phq_cat",
  type = "response"
)

interaction_treatment_low_phq <- interaction_treatment_effects |>
  filter(phq_cat == "< 10")
interaction_treatment_high_phq <- interaction_treatment_effects |>
  filter(phq_cat == ">= 10")

interaction_log_odds_effects <- avg_comparisons(
  mod_interaction,
  variables = list(trt = c("waitlist", "therapist-guided")),
  by = "phq_cat",
  type = "link"
) |>
  mutate(
    odds_ratio = exp(estimate),
    odds_ratio_low = exp(conf.low),
    odds_ratio_high = exp(conf.high)
  )

interaction_odds_ratio_low_phq <- interaction_log_odds_effects |>
  filter(phq_cat == "< 10")
interaction_odds_ratio_high_phq <- interaction_log_odds_effects |>
  filter(phq_cat == ">= 10")

summary(mod_interaction)
interaction_probabilities
interaction_treatment_effects
```

::: {.callout-tip collapse="true"}
## Model interpretation

Within the baseline PHQ-9 `< 10` category, therapist-guided treatment has an estimated probability of a low post-treatment LSAS score that is **`r if (interaction_treatment_low_phq$estimate[1] >= 0) "higher" else "lower"` by `r sprintf("%.1f", 100 * abs(interaction_treatment_low_phq$estimate[1]))` percentage points** than waitlist (95% CI **`r sprintf("%.1f", 100 * interaction_treatment_low_phq$conf.low[1])` to `r sprintf("%.1f", 100 * interaction_treatment_low_phq$conf.high[1])` percentage points**). Within the baseline PHQ-9 `>= 10` category, the corresponding probability is **`r if (interaction_treatment_high_phq$estimate[1] >= 0) "higher" else "lower"` by `r sprintf("%.1f", 100 * abs(interaction_treatment_high_phq$estimate[1]))` percentage points** (95% CI **`r sprintf("%.1f", 100 * interaction_treatment_high_phq$conf.low[1])` to `r sprintf("%.1f", 100 * interaction_treatment_high_phq$conf.high[1])` percentage points**).

The associated therapist-guided-versus-waitlist odds ratios are **`r sprintf("%.2f", interaction_odds_ratio_low_phq$odds_ratio[1])`** (95% CI **`r sprintf("%.2f", interaction_odds_ratio_low_phq$odds_ratio_low[1])` to `r sprintf("%.2f", interaction_odds_ratio_low_phq$odds_ratio_high[1])`**) for PHQ-9 `< 10` and **`r sprintf("%.2f", interaction_odds_ratio_high_phq$odds_ratio[1])`** (95% CI **`r sprintf("%.2f", interaction_odds_ratio_high_phq$odds_ratio_low[1])` to `r sprintf("%.2f", interaction_odds_ratio_high_phq$odds_ratio_high[1])`**) for PHQ-9 `>= 10`. They compare odds, not probabilities. Use the conditional probability differences above—not these odds ratios—to describe the clinical contrast.
:::

```{r}
#| label: plot-interaction-probabilities
#| fig-cap: "Interaction-model estimated probabilities of a low post-treatment LSAS score. Error bars are 95% confidence intervals."
interaction_probabilities |>
  ggplot(aes(x = trt, y = estimate, group = phq_cat, colour = phq_cat)) +
  geom_point(position = position_dodge(width = 0.1)) +
  geom_line(position = position_dodge(width = 0.1)) +
  geom_errorbar(
    aes(ymin = conf.low, ymax = conf.high),
    width = 0.1,
    position = position_dodge(width = 0.1)
  ) +
  labs(
    title = "Interaction logistic model",
    subtitle = "Treatment probabilities may differ by baseline PHQ-9 category",
    x = "Treatment group",
    y = "Estimated probability of low post-treatment LSAS (95% CI)",
    colour = "Baseline PHQ-9"
  ) +
  theme_minimal()
```

```{r}
#| label: plot-conditional-treatment-effects
#| fig-cap: "Conditional therapist-guided-minus-waitlist probability differences from the interaction model. Error bars are 95% confidence intervals."
plot_comparisons(
  mod_interaction,
  variables = list(trt = c("waitlist", "therapist-guided")),
  by = "phq_cat",
  type = "response"
) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  labs(
    title = "Conditional treatment effects on the probability scale",
    x = "Baseline PHQ-9 category",
    y = "Therapist-guided minus waitlist probability difference (95% CI)"
  ) +
  theme_minimal()
```

# Interpreting the interaction

An interaction model can describe different conditional estimates, but it does not automatically establish a clinically meaningful subgroup. Before presenting a subgroup finding, ask:

1. Was the moderator and interaction planned before looking at the data?
2. Are the conditional probability differences precise enough to distinguish meaningful patterns from sampling variation?
3. Does the interaction plot show a consistent, clinically interpretable pattern rather than an isolated coefficient or p-value?
4. Would the conclusion be different if PHQ-9 were modelled continuously? See [logistic regression with a binary and numeric predictor](logistic-binary-numeric-predictors.qmd).

The interaction coefficient is a difference of **log-odds contrasts**. It is not itself the difference between the two probability differences. `avg_comparisons()` gives the quantities we want to communicate: therapist-guided-minus-waitlist probability differences within the explicitly stated PHQ-9 categories.

# Summary

In this chapter you learned to:

- derive `phq_cat` explicitly as baseline PHQ-9 `< 10` versus `>= 10`;
- fit additive and `trt * phq_cat` binary logistic models with `stats::glm(..., family = binomial)`;
- report treatment contrasts as conditional probability differences in percentage points with 95% CIs;
- distinguish a conditional odds ratio from a probability difference; and
- use predicted-probability and conditional-effect plots to inspect a proposed subgroup pattern.
