---
title: "Logistic regression with a binary and numeric predictor"
format:
  html:
    df-print: kable
toc: true
toc-depth: 3
number-sections: true
editor: 
  markdown: 
    wrap: 72
---

# A continuous moderator

In [logistic regression with two binary
predictors](logistic-two-binary-predictors.qmd), baseline depression was
grouped as below or at least 10 on the PHQ-9. That can be useful for a
clearly defined clinical decision, but it also turns different scores
into the same group. Here we retain the observed continuous
`phq9_screen` score and ask:

> How does the estimated probability of treatment response differ
> between therapist-guided treatment and waitlist at different baseline
> PHQ-9 scores?

We will make that question concrete at PHQ-9 scores of 5, 10, and 15.
They represent low, moderate, and higher values within the observed
range; they are not diagnostic categories. Our primary result is a
**therapist-guided minus waitlist probability difference**, expressed in
percentage points with a 95% CI.

A logistic-regression interaction is an interaction on the **log-odds
scale**. That detail matters: even an additive logistic model can have
probability differences that vary with PHQ-9, because the
logit-to-probability transformation is nonlinear. We therefore calculate
contrasts on the probability scale rather than reading their shape from
a coefficient table.

# Packages and analytic data

```{r}
#| label: setup
#| message: false
#| warning: false
library(tidyverse)
library(here)
library(marginaleffects)

options(
  marginaleffects_print_omit = c("s.value"),
  marginaleffects_print_style = ifelse(interactive(), "summary", "tinytable")
)
```

```{r}
#| label: make-analysis-data
#| message: false
#| warning: false
df_binary <- read_csv2(here("data", "steps_clean.csv")) |>
  filter(!is.na(lsas_screen), !is.na(lsas_post)) |>
  mutate(
    # The second factor level is the event modeled by binomial glm().
    lsas_response = factor(
      if_else(
        (lsas_screen - lsas_post) / lsas_screen >= 0.30,
        "Response",
        "No response"
      ),
      levels = c("No response", "Response")
    ),
    trt = factor(
      trt,
      levels = c("waitlist", "therapist-guided", "self-guided")
    )
  ) |>
  filter(trt %in% c("waitlist", "therapist-guided")) |>
  drop_na(lsas_response, phq9_screen) |>
  droplevels()

phq_values <- c(5, 10, 15)

df_binary |>
  summarise(
    participants = n(),
    minimum_phq9 = min(phq9_screen),
    maximum_phq9 = max(phq9_screen)
  )
```

`Response` means a reduction of at least 30% from baseline LSAS. This is
a response outcome, not a diagnostic or remission definition. Waitlist
is the treatment reference group, and `Response` is the event whose
probability is modeled. We restrict this chapter to the waitlist and
therapist-guided arms so that every treatment contrast has one clear
direction.

# Additive logistic model

## What does “additive” mean here?

The additive model has a treatment term and a PHQ-9 term, but no product
term:

$$
\operatorname{logit}\{P(Y_i = \text{Response} \mid trt_i, PHQ9_i)\} =
\beta_0 +
\beta_1\mathbf{1}\{trt_i = \text{therapist-guided}\} +
\beta_2 PHQ9_i.
$$

The therapist-guided versus waitlist **log-odds** difference is the same
at every PHQ-9 score in this model. Equivalently, the treatment odds
ratio is constant across PHQ-9. That is not the same as saying the
probability difference is constant: a fixed change in log-odds maps to
different probability changes near 0, 0.5, and 1.

```{r}
#| label: fit-additive-logit
#| message: false
#| warning: false
mod_additive <- stats::glm(
  lsas_response ~ trt + phq9_screen,
  family = binomial,
  data = df_binary
)

summary(mod_additive)
```

We can exponentiate the coefficients to express them on the odds-ratio
scale. The treatment coefficient gives the odds ratio for
therapist-guided treatment versus waitlist, adjusted for baseline PHQ-9.
Because this model does not include an interaction, this treatment odds
ratio is assumed to be the same at every baseline PHQ-9 score.

For PHQ-9, the coefficient represents the change in log-odds associated
with a **one-point increase** in baseline PHQ-9. Exponentiating this
coefficient therefore gives the odds ratio associated with a one-point
higher PHQ-9 score, adjusted for treatment group.

```{r}
#| label: fit-additive-logit-manual
#| message: false
#| warning: false
b_additive <- coef(mod_additive)
exp(b_additive[["trttherapist-guided"]]) # OR therapist-guided vs waitlist
exp(b_additive[["phq9_screen"]]) # OR for a one-point increase in PHQ-9
```

Thus, the estimated treatment OR is approximately
**`r sprintf("%.2f", exp(b_additive[["trttherapist-guided"]]))`**, while
the OR associated with a one-point increase in PHQ-9 is approximately
**`r sprintf("%.2f", exp(b_additive[["phq9_screen"]]))`**.

As stated before, odds ratios can be difficult to translate into the
actual probability of treatment response. We can therefore use the
fitted model to estimate the treatment contrast directly on the
**probability scale**. `avg_comparisons()` predicts each participant's
probability of response under therapist-guided treatment and under
waitlist, calculates the difference between these probabilities, and
then averages these differences across participants.

```{r}
#| label: fit-additive-logit-probabilities
#| message: false
#| warning: false
additive_trt_difference <- avg_comparisons(
  mod_additive,
  variables = list(trt = c("waitlist", "therapist-guided")),
  type = "response"
)
additive_trt_difference
```

::: {.callout-tip collapse="true"}
## Model interpretation

Averaged over the observed baseline PHQ-9 distribution, therapist-guided
treatment is estimated to change the probability of treatment response
by
**`r sprintf("%.1f", 100 * additive_trt_difference$estimate[1])`
percentage points** compared with waitlist (95% CI
**`r sprintf("%.1f", 100 * additive_trt_difference$conf.low[1])`** to
**`r sprintf("%.1f", 100 * additive_trt_difference$conf.high[1])`
percentage points**).

This is therapist-guided minus waitlist; a positive value favors a
higher estimated probability of response with therapist guidance. It is
a population-average probability contrast over this analysis sample,
conditional on the additive model—not a coefficient interpreted as a
percentage-point change.
:::

```{r}
#| label: plot-additive-probabilities
#| fig-cap: "Additive-model estimated probabilities of treatment response, with 95% confidence intervals."
plot_predictions(
  mod_additive,
  condition = c("phq9_screen", "trt")
) +
  labs(
    title = "Additive logistic model",
    subtitle = "Estimated probabilities with 95% confidence intervals",
    x = "Baseline PHQ-9 score",
    y = "Probability of treatment response",
    colour = "Treatment group",
    fill = "Treatment group"
  ) +
  theme_minimal()
```

The curves need not be parallel on the probability scale. The additive
model makes its parallel-lines assumption on the log-odds scale, not on
this plot's probability scale.

# Interaction model

## Let the treatment association vary with PHQ-9

The interaction model adds a product term:

$$
\begin{aligned}
\operatorname{logit}\{P(Y_i = \text{Response} \mid trt_i, PHQ9_i)\} =
&\ \beta_0 + \beta_1\mathbf{1}\{trt_i = \text{therapist-guided}\} + \\
&\ \beta_2 PHQ9_i +
\beta_3\mathbf{1}\{trt_i = \text{therapist-guided}\} \times PHQ9_i.
\end{aligned}
$$

Now the treatment log-odds difference can change with PHQ-9. The
treatment coefficient in `summary()` is the difference at PHQ-9 = 0,
which is not a useful stand-alone clinical summary here. We instead
estimate probability differences at prespecified, observed-range scores.

```{r}
#| label: fit-interaction-logit
#| message: false
#| warning: false
mod_interaction <- stats::glm(
  lsas_response ~ trt * phq9_screen,
  family = binomial,
  data = df_binary
)

summary(mod_interaction)

# Store the coefficients for the calculations below
b_interaction <- coef(mod_interaction)
b_trt <- b_interaction[["trttherapist-guided"]]
b_trt_phq <- b_interaction[["trttherapist-guided:phq9_screen"]]
```

With an interaction in the model, the association between treatment and
response is allowed to vary with baseline PHQ-9. This changes how we
interpret the coefficients.

The treatment coefficient, `r sprintf("%.2f", b_trt)`, is the
therapist-guided versus waitlist difference in log-odds when PHQ-9 = 0.
Exponentiating it gives an odds ratio of approximately
`r sprintf("%.1f", exp(b_trt))`, but this is the treatment odds ratio
specifically at PHQ-9 = 0 and is therefore not particularly informative
for our clinical question.

The interaction coefficient tells us how the treatment difference on the
log-odds scale changes with each one-point increase in PHQ-9. Here, the
interaction coefficient is `r sprintf("%.3f", b_trt_phq)`. This means
that for each one-point increase in PHQ-9, the therapist-guided versus
waitlist log-odds difference
`r if (b_trt_phq < 0) "decreases" else "increases"` by
`r sprintf("%.3f", abs(b_trt_phq))`.

We can exponentiate the interaction coefficient:

```{r}
exp(b_trt_phq)
```

This gives approximately **`r sprintf("%.2f", exp(b_trt_phq))`**. Thus,
for every one-point increase in PHQ-9, the treatment odds ratio is
multiplied by approximately `r sprintf("%.2f", exp(b_trt_phq))`. For
example, the treatment odds ratio at PHQ-9 = 10 is:

```{r}
exp(b_trt + b_trt_phq * 10)
```

However, interpreting interactions through odds ratios quickly becomes
cumbersome. It is usually easier to ask our substantive question
directly: **how much higher or lower is the estimated probability of
treatment response with therapist-guided treatment at different PHQ-9
scores?**

We can therefore estimate the therapist-guided versus waitlist
probability difference at selected PHQ-9 values:

```{r}
#| label: fit-interaction-logit-probabilities
#| message: false
#| warning: false

conditional_trt_difference <- avg_comparisons(
  mod_interaction,
  variables = list(trt = c("waitlist", "therapist-guided")),
  newdata = datagrid(phq9_screen = phq_values),
  by = "phq9_screen",
  type = "response"
)
conditional_trt_difference
```

::: {.callout-tip collapse="true"}
## Model interpretation

At PHQ-9 = 5, therapist-guided minus waitlist is an estimated
**`r sprintf("%.1f", 100 * conditional_trt_difference$estimate[conditional_trt_difference$phq9_screen == 5])`
percentage-point** difference in the probability of treatment response
(95% CI
**`r sprintf("%.1f", 100 * conditional_trt_difference$conf.low[conditional_trt_difference$phq9_screen == 5])`**
to
**`r sprintf("%.1f", 100 * conditional_trt_difference$conf.high[conditional_trt_difference$phq9_screen == 5])`**).

At PHQ-9 = 10, the corresponding difference is
**`r sprintf("%.1f", 100 * conditional_trt_difference$estimate[conditional_trt_difference$phq9_screen == 10])`
percentage points** (95% CI
**`r sprintf("%.1f", 100 * conditional_trt_difference$conf.low[conditional_trt_difference$phq9_screen == 10])`**
to
**`r sprintf("%.1f", 100 * conditional_trt_difference$conf.high[conditional_trt_difference$phq9_screen == 10])`**);
at PHQ-9 = 15 it is
**`r sprintf("%.1f", 100 * conditional_trt_difference$estimate[conditional_trt_difference$phq9_screen == 15])`
percentage points** (95% CI
**`r sprintf("%.1f", 100 * conditional_trt_difference$conf.low[conditional_trt_difference$phq9_screen == 15])`**
to
**`r sprintf("%.1f", 100 * conditional_trt_difference$conf.high[conditional_trt_difference$phq9_screen == 15])`**).

Each contrast conditions on the displayed PHQ-9 value. Positive values
mean a higher modeled probability with therapist-guided treatment. The
estimates and intervals, rather than the interaction coefficient alone,
describe whether the clinically relevant probability difference changes
across PHQ-9.
:::

```{r}
#| label: plot-interaction-probabilities
#| fig-cap: "Interaction-model estimated probabilities of treatment response, with 95% confidence intervals."
plot_predictions(
  mod_interaction,
  condition = c("phq9_screen", "trt")
) +
  labs(
    title = "Interaction logistic model",
    subtitle = "Estimated probabilities with 95% confidence intervals",
    x = "Baseline PHQ-9 score",
    y = "Probability of treatment response",
    colour = "Treatment group",
    fill = "Treatment group"
  ) +
  theme_minimal()
```

## A curve of the conditional probability difference

The probability curves are useful, but they leave the reader to visually
subtract two uncertain quantities. A direct contrast plot answers the
treatment question more directly.

```{r}
#| label: plot-conditional-treatment-difference
#| fig-cap: "Conditional therapist-guided-minus-waitlist probability difference across baseline PHQ-9, with 95% confidence intervals."
plot_comparisons(
  mod_interaction,
  variables = list(trt = c("waitlist", "therapist-guided")),
  condition = "phq9_screen",
  type = "response"
) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  labs(
    title = "Conditional treatment probability difference",
    subtitle = "Therapist-guided minus waitlist; 95% confidence intervals",
    x = "Baseline PHQ-9 score",
    y = "Probability difference"
  ) +
  scale_y_continuous(labels = scales::label_percent(accuracy = 1)) +
  theme_minimal()
```

The dashed zero line marks no modeled probability difference. Wide
intervals at the edge of the observed PHQ-9 range are a reminder not to
make subgroup claims from the curve's point estimates alone.

# Conditional and population-average treatment contrasts

A conditional contrast answers a question such as “what is the modeled
difference at PHQ-9 = 10?” A population-average contrast answers a
different question: “what is the modeled treatment difference after
averaging each participant's counterfactual predictions over the
observed PHQ-9 distribution?”

```{r}
#| label: interaction-marginal-treatment-difference
marginal_trt_difference <- avg_comparisons(
  mod_interaction,
  variables = list(trt = c("waitlist", "therapist-guided")),
  type = "response"
)
marginal_trt_difference
```

::: {.callout-tip collapse="true"}
## Model interpretation

Averaged over the observed PHQ-9 distribution, the interaction model
estimates that therapist-guided treatment changes the probability of
treatment response by
**`r sprintf("%.1f", 100 * marginal_trt_difference$estimate[1])`
percentage points** compared with waitlist (95% CI
**`r sprintf("%.1f", 100 * marginal_trt_difference$conf.low[1])`** to
**`r sprintf("%.1f", 100 * marginal_trt_difference$conf.high[1])`
percentage points**).

This population-average contrast is not the effect at a particular PHQ-9
score. It averages conditional contrasts over the PHQ-9 values
represented in this sample, so it can change if the target population's
baseline PHQ-9 distribution changes.
:::

# Non-collapsibility of the odds ratio

Adjusting for a covariate can change a treatment estimate for many reasons, e.g. by dealing with confounding.
In a randomized trial, adjusting for a prognostic baseline covariate is commonly done to increase precision [@FDA2023Covariate; @Morris2022CovariateAdjustment].
In a linear model of post-treatment LSAS, such adjustment leaves the treatment coefficient unchanged on average.
Odds ratios behave differently: when the treatment has an effect, adjusting for a prognostic baseline covariate moves the odds ratio away from 1, on average, even in a perfectly balanced trial [@Robinson1991Covariate; @Daniel2021NonCollapsible].
This property is called non-collapsibility.

## Collapsible and non-collapsible effect measures

An effect measure is *collapsible* if the effect in a whole population is a weighted average of the effects in its subgroups.
The population effect then always lies between the smallest and largest subgroup effects, which is what most of us would expect of any effect measure.
Risk differences and risk ratios behave this way.
The odds ratio does not: the odds ratio in the whole population can lie outside the range of the subgroup odds ratios, even when nothing is confounded [@Hernan2020WhatIf; @Greenland1999Collapsibility].

The following made-up trial, adapted from @Hernan2020WhatIf, shows what this means.
Treatment is randomized among 400 participants, half with mild and half with severe symptoms, and severity strongly predicts response.

```{r}
#| label: collapsibility-example
df_collapse <- tibble(
  severity = rep(c("mild", "severe"), each = 2),
  trt = rep(c("control", "treatment"), times = 2),
  n = 100,
  responders = c(40, 80, 20, 60)
)

# Risk difference, risk ratio, and odds ratio from the response proportions
effect_measures <- function(data) {
  data |>
    summarise(p = sum(responders) / sum(n), .by = trt) |>
    pivot_wider(names_from = trt, values_from = p) |>
    transmute(
      risk_difference = treatment - control,
      risk_ratio = treatment / control,
      odds_ratio = (treatment / (1 - treatment)) / (control / (1 - control))
    )
}

collapse_measures <- list(
  Mild = filter(df_collapse, severity == "mild"),
  Severe = filter(df_collapse, severity == "severe"),
  `Whole trial` = df_collapse
) |>
  map(effect_measures) |>
  list_rbind(names_to = "group")
collapse_measures
```

In both severity groups, treatment increases the probability of response by `r sprintf("%.0f", 100 * collapse_measures$risk_difference[1])` percentage points, and so it does in the whole trial.
The risk ratio is `r sprintf("%.0f", collapse_measures$risk_ratio[1])` in the mild group and `r sprintf("%.0f", collapse_measures$risk_ratio[2])` in the severe group, and `r sprintf("%.2f", collapse_measures$risk_ratio[3])` in the whole trial, in between the two.
The odds ratio is `r sprintf("%.0f", collapse_measures$odds_ratio[1])` in both groups, yet `r sprintf("%.2f", collapse_measures$odds_ratio[3])` in the whole trial: smaller than in either half.
Severity is not a confounder, because treatment was randomized. And because the made-up trial is exactly balanced, with 100 participants in each cell, chance imbalance plays no role either, so the difference between the odds ratios is due to non-collapsibility alone.
The reason lies in how the whole-trial proportions arise.
Each whole-trial response proportion is an average of the subgroup proportions, because each severity group makes up half of each arm.
The whole-trial risk difference is therefore the average of the subgroup risk differences.
The whole-trial risk ratio is a weighted average of the subgroup risk ratios, so it lies between them.
Odds are a nonlinear function of probability, so the whole-trial odds are not averages of the subgroup odds, and the whole-trial odds ratio need not lie between the subgroup odds ratios; this is the point made in [logistic regression with a numeric predictor](logistic-one-numeric-predictor.qmd).

## Conditional and marginal odds ratios in regression models

Regression models show the same thing.
Adjusting for severity gives the odds ratio within severity groups, whereas the unadjusted model gives the odds ratio for the whole trial:

```{r}
#| label: collapsibility-models
# One row per participant
df_collapse_long <- df_collapse |>
  reframe(
    response = rep(c(1, 0), c(responders, n - responders)),
    .by = c(severity, trt)
  ) |>
  mutate(trt = factor(trt, levels = c("control", "treatment")))

mod_collapse_unadjusted <- glm(
  response ~ trt,
  family = binomial,
  data = df_collapse_long
)
mod_collapse_adjusted <- glm(
  response ~ trt + severity,
  family = binomial,
  data = df_collapse_long
)

# Odds ratio from the coefficient, standardized odds ratio, and standardized
# risk difference for each model
summarise_collapse <- function(mod) {
  trt_contrast <- list(trt = c("control", "treatment"))
  tibble(
    or_coefficient = exp(coef(mod)[["trttreatment"]]),
    or_standardized = avg_comparisons(
      mod,
      variables = trt_contrast,
      comparison = "lnoravg",
      transform = exp
    )$estimate,
    rd_standardized = avg_comparisons(mod, variables = trt_contrast)$estimate
  )
}

collapse_models <- list(
  Unadjusted = mod_collapse_unadjusted,
  `Adjusted for severity` = mod_collapse_adjusted
) |>
  map(summarise_collapse) |>
  list_rbind(names_to = "model")
collapse_models
```

::: {.callout-tip collapse="true"}
## Model interpretation

The odds ratio from the treatment coefficient is **`r sprintf("%.2f", collapse_models$or_coefficient[1])`** in the unadjusted model and **`r sprintf("%.2f", collapse_models$or_coefficient[2])`** in the model adjusted for severity.
Standardizing the adjusted model gives back the whole-trial odds ratio, **`r sprintf("%.2f", collapse_models$or_standardized[2])`**.
The standardized risk difference is **`r sprintf("%.0f", 100 * collapse_models$rd_standardized[1])` percentage points** in both models.
:::

Both models are correct; they estimate different odds ratios.
The conditional odds ratio describes the treatment effect among participants with the same severity, whereas the marginal odds ratio describes the effect averaged over the whole trial.
Standardizing the adjusted model recovers the marginal odds ratio, and the standardized risk difference does not depend on whether severity is in the model, because risk differences are collapsible.

# Summary

- An additive logistic model assumes a common treatment **log-odds**
  difference, not necessarily a constant probability difference.
- An interaction model lets the treatment association vary with
  continuous baseline PHQ-9.
- `avg_comparisons(..., type = "response")` supplies conditional and
  population-average treatment contrasts on the probability scale.
- A contrast curve communicates both the estimated pattern and its
  uncertainty.
- Odds ratios describe ratios of odds. They are not percentage-point
  changes in probability.
- Odds ratios are non-collapsible: adjusting for a prognostic covariate
  changes the odds ratio even without confounding, whereas standardized
  risk differences keep their meaning.
