---
title: "Logistic regression with a binary predictor"
format:
  html:
    df-print: kable
toc: true
toc-depth: 3
number-sections: true
editor: 
  markdown: 
    wrap: 72
---

# Introduction

This notebook extends the treatment example from [logistic-regression
foundations](logistic-intro.qmd). We focus on the two arms that make a
binary treatment predictor: `waitlist` and `therapist-guided`.

Our question is:

> Among participants in these two arms, how does the estimated
> probability of treatment response differ for therapist-guided
> treatment versus waitlist?

The target is the **therapist-guided minus waitlist probability
difference**. We will use the corresponding odds ratio to read
logistic-regression output, but the probability difference is the
primary clinical result.

# Packages and data

```{r}
#| label: setup
#| message: false
#| warning: false
library(tidyverse)
library(here)
library(broom)
library(marginaleffects)

options(
  marginaleffects_print_omit = c("s.value"),
  marginaleffects_print_style = ifelse(interactive(), "summary", "tinytable")
)
```

```{r}
#| label: load-binary-treatment-data
#| message: false
#| warning: false
df_binary <- read_csv2(here("..", "..", "data", "steps_clean.csv")) |>
  filter(!is.na(lsas_screen), !is.na(lsas_post)) |>
  mutate(
    # The second level is the event modelled by binomial glm().
    lsas_response = factor(
      if_else(
        (lsas_screen - lsas_post) / lsas_screen >= 0.30,
        "Response",
        "No response"
      ),
      levels = c("No response", "Response")
    ),
    trt = factor(
      trt,
      levels = c("waitlist", "therapist-guided", "self-guided")
    )
  ) |>
  filter(trt %in% c("waitlist", "therapist-guided")) |>
  droplevels()

df_binary |>
  count(trt, lsas_response, .drop = FALSE)
```

`Response` means that LSAS decreases by at least 30% from baseline, as
in [logistic-regression foundations](logistic-intro.qmd).

We explicitly make waitlist the reference group. The treatment contrast
shown by the coefficient will therefore be log-odds of response for the
therapist-guided group minus the log-odds for treatment response in the
waitlist group.

```{r}
#| label: inspect-binary-treatment-coding
levels(df_binary$trt)
contrasts(df_binary$trt)
```

# A logistic model for two treatment groups

For participant $i$, let $Y_i$ indicate treatment response. The model is

$$
\operatorname{logit}\left[P(Y_i = \text{Response} \mid trt_i)\right] =
\beta_0 + \beta_1\,\mathbf{1}\{trt_i = \text{therapist-guided}\}.
$$

Here, $\beta_0$ is the log-odds of response in the waitlist group, and
$\beta_1$ is the therapist-guided-versus-waitlist log-odds difference.
These are the quantities that `glm()` estimates directly. They are not
yet the most clinically intelligible quantities.

## Fit and interpret the model

We fit the model using the `glm()` function, with the
`family = binomial` to tell `R` that this is a logistic model. We can
then use the `summary()` function to see the model coefficients.

```{r}
#| label: fit-binary-treatment-logit
#| message: false
#| warning: false
mod_binary_trt <- stats::glm(
  lsas_response ~ trt, # model formula
  family = binomial, # using the binomial family to get a logistic regression
  data = df_binary # using our dataset df_binary
)

summary(mod_binary_trt)
```

The output from `summary()` shows us that the log-odds in the waitlist
group is -2.16 and that the log-odds in the therapist-guided group is
2.46 higher than that of the waitlist (i.e. -2.16 + 2.46).

By exponentiating the linear predictors (added coefficients), **we get
the odds for each group.** By exponentiating only the coefficient for
the therapist-guided group (2.46; the difference in log-odds between the
therapist-guided and the waitlist group), we get the **odds ratio of
response in the therapist-guided vs. the waitlist group**

```{r}
#| label: interpret-binary-treatment-logit-manual
exp(-2.16) # odds of response in waitlist group
exp(-2.16 + 2.46) # odds of response in therapist-guided group
exp(2.46) # odds ratio of response in the therapist-guided group vs. the waitlist
```

To go from the linear predictors of the log-odds to the probability, we
can use the inverse logit function:

$$\operatorname{logit}^{-1}(x) = \frac{e^x}{1 + e^x}$$

| Treatment group | Linear predictor | Model-derived probability |
|--------------------|-------------------------:|-------------------------:|
| waitlist | $\beta_0$ | $\operatorname{logit}^{-1}(\beta_0)$ |
| therapist-guided | $\beta_0 + \beta_1$ | $\operatorname{logit}^{-1}(\beta_0 + \beta_1)$ |

```{r}
#| label: predicted-probability-binary-treatment-logit-manual
exp(-2.16) / (1 + exp(-2.16)) # probability of response in waitlist group
exp(-2.16 + 2.46) / (1 + exp(-2.16 + 2.46)) # probability of response in therapist group
```

## Get probabilities and the probability difference from the model

You can also get these predicted probabilities using the
`marginaleffects` package.

```{r}
#| label: show-binary-treatment-estimates

p_binary_trt <- avg_predictions(
  mod_binary_trt,
  by = "trt",
  type = "response"
)

diff_binary_trt <- avg_comparisons(
  mod_binary_trt,
  variables = list(trt = c("waitlist", "therapist-guided")),
  type = "response"
)

or_binary_trt <- tidy(
  mod_binary_trt,
  conf.int = TRUE,
  exponentiate = TRUE
) |>
  filter(term == "trttherapist-guided")

p_binary_trt

diff_binary_trt
```

The estimated probability difference is calculated after transforming
the model to the response scale. It answers a different question from
the printed log-odds coefficient: how many percentage points higher or
lower is the modelled probability in one group than the other?

::: {.callout-tip collapse="true"}
## Model interpretation

The estimated probability of treatment response is
**`r sprintf("%.1f", 100 * p_binary_trt$estimate[p_binary_trt$trt == "waitlist"])`%**
for waitlist (95% CI
**`r sprintf("%.1f", 100 * p_binary_trt$conf.low[p_binary_trt$trt == "waitlist"])`%**
to
**`r sprintf("%.1f", 100 * p_binary_trt$conf.high[p_binary_trt$trt == "waitlist"])`%**)
and
**`r sprintf("%.1f", 100 * p_binary_trt$estimate[p_binary_trt$trt == "therapist-guided"])`%**
for therapist-guided treatment (95% CI
**`r sprintf("%.1f", 100 * p_binary_trt$conf.low[p_binary_trt$trt == "therapist-guided"])`%**
to
**`r sprintf("%.1f", 100 * p_binary_trt$conf.high[p_binary_trt$trt == "therapist-guided"])`%**).

Compared with waitlist, therapist-guided treatment
**`r ifelse(diff_binary_trt$estimate[1] >= 0, "increases", "decreases")`**
the estimated probability by
**`r sprintf("%.1f", abs(100 * diff_binary_trt$estimate[1]))` percentage
points** (therapist-guided minus waitlist =
**`r sprintf("%.1f", 100 * diff_binary_trt$estimate[1])` percentage
points**; 95% CI
**`r sprintf("%.1f", 100 * diff_binary_trt$conf.low[1])`** to
**`r sprintf("%.1f", 100 * diff_binary_trt$conf.high[1])` percentage
points**). This is an unadjusted contrast within the restricted two-arm
sample.

The associated therapist-guided-versus-waitlist **odds ratio** is
**`r sprintf("%.2f", or_binary_trt$estimate[1])`** (95% CI
**`r sprintf("%.2f", or_binary_trt$conf.low[1])`** to
**`r sprintf("%.2f", or_binary_trt$conf.high[1])`**). It is a ratio of
the odds of response, not a probability ratio and not a
`r sprintf("%.1f", 100 * diff_binary_trt$estimate[1])`-percentage-point
change.
:::

## Visualise the group probabilities

```{r}
#| label: plot-binary-treatment-probabilities
plot_predictions(mod_binary_trt, by = "trt", type = "response") +
  labs(
    title = "Estimated probability of treatment response",
    subtitle = "Waitlist versus therapist-guided treatment; 95% confidence intervals",
    x = "Treatment group",
    y = "Probability of treatment response"
  ) +
  theme_minimal()
```

# Log-odds and odds ratios

The logistic model is linear on the **log-odds** scale. Exponentiating
$\beta_1$ gives the therapist-guided-versus-waitlist odds ratio:

$$
\exp(\beta_1) =
\frac{\text{odds of response with therapist-guided treatment}}
{\text{odds of response with waitlist}}.
$$

```{r}
#| label: show-binary-treatment-odds-ratio
or_binary_trt
```

An odds ratio larger than 1 means higher *odds* of response in the
therapist-guided group. It cannot be read as a percentage-point
increase, because the same odds ratio implies different probability
differences at different baseline probabilities. Report the
model-derived probabilities and their probability difference when the
clinical question concerns chances of an outcome.

# Summary

In this chapter you learned to:

- restrict the STePS data to a binary treatment comparison and set
  waitlist as the reference category;
- fit a binary logistic model with `stats::glm(..., family = binomial)`;
- report treatment-specific probabilities and the
  therapist-guided-minus-waitlist probability difference with 95% CIs;
  and
- interpret the associated odds ratio as a ratio of odds rather than a
  probability difference.
