---
title: "Logistic regression foundations"
format:
  html:
    df-print: kable
toc: true
toc-depth: 3
number-sections: true
---

# Why logistic regression?

Course 2 used linear and quantile regression to model continuous outcomes such as post-treatment LSAS scores. In many psychiatric research questions, the outcome is instead binary:

- remission versus non-remission;
- response versus non-response;
- attendance versus dropout; or
- an adverse event versus no adverse event.

Logistic regression and its multinomial extension are helpful tools when you want to analyse a binary or categorical outcome. This chapter uses a binary post-treatment social-anxiety outcome. Our main question is:

> How does the probability of treatment response differ between treatment groups?

But first, some setup.

# Packages and data

```{r}
#| label: setup
#| message: false
#| warning: false
library(tidyverse)
library(here)
library(broom)
library(marginaleffects)
library(tinytable)

options(
  marginaleffects_print_omit = c("s.value"),
  marginaleffects_print_style = ifelse(interactive(), "summary", "tinytable")
)
```

```{r}
#| label: load-data
#| message: false
#| warning: false
df_clean <- read_csv2(here("..", "..", "data", "steps_clean.csv")) |>
  filter(!is.na(lsas_screen), !is.na(lsas_post)) |>
  mutate(
    # The second factor level is the event modelled by binomial glm().
    lsas_response = factor(
      if_else((lsas_screen - lsas_post) / lsas_screen >= 0.30, "Response", "No response"),
      levels = c("No response", "Response")
    ),
    trt = factor(
      trt,
      levels = c("waitlist", "therapist-guided", "self-guided")
    )
  )

df_clean |>
  count(lsas_response, .drop = FALSE)
```

The binary outcome is treatment response:

``` r
lsas_response = "Response"    if LSAS decreases by at least 30%
lsas_response = "No response" otherwise
```

This follows the practical treatment-follow-up guidance for LSAS-SR, which describes improvement as an approximately 30% reduction in total score [Nationell formulärsamling](https://patientenkat.se/nationellpatientenkat/nationellformularsamling/sokformularbeskrivningar/lsassrsocialangest.92403.html). In a German treated-SAD sample, the empirically derived LSAS-SR response cut-point was 28% [@vonGlischinski2018LSAS]. We use the rounded 30% definition in this course.

# Defining the binary outcome

The LSAS-SR is often used to screen for social anxiety disorder. The original clinical LSAS study found total-score cut-points of 30 for social anxiety disorder and 60 for the then-defined generalised subtype [@Mennin2002LSAS]. A study of the self-report LSAS-SR found the same cut-points had a useful sensitivity/specificity balance [@Rytwinski2009LSASSR].

Thresholds can serve different purposes. In the same treated German sample, an LSAS-SR score of 35 was an empirically derived remission cut-point [@vonGlischinski2018LSAS]. Response, remission, and diagnostic screening are different outcomes and should not be used interchangeably.

::: callout-caution
## A response definition is an analysis decision

The 30% response definition is useful for this course, but it is not a diagnosis or a remission definition. Before creating a binary clinical outcome in your own work, specify the purpose and identify a defensible external definition.
:::

# Intercept-only logistic regression

We first ignore treatment group and estimate one probability: the probability that a participant has treatment response at post-treatment.

The model is:

$$
\operatorname{logit}\left[P(Y_i = \text{Response})\right] = \beta_0,
$$

where the logit transforms a probability $p$ into log-odds:

$$
\operatorname{logit}(p) = \log\left(\frac{p}{1-p}\right).
$$

```{r}
#| label: fit-intercept-only
mod_intercept <- glm(
  lsas_response ~ 1,
  family = binomial,
  data = df_clean
)

summary(mod_intercept)
```

The intercept is on the log-odds scale, so it is not the clearest clinical summary. If we exponentiate the estimate, we get it on the odds scale. This is the same as dividing the the number of repsonders (n=57) by the the number of non-responders (n=112).

```{r}
#| label: manual-log-odds
exp(-0.6754) 
57/112
```

The odds is still not the clearest summary, usually the probability is more easily interpreted. We can use `marginaleffects` to obtain the estimated probability and its confidence interval.

```{r}
#| label: intercept-probability
p_overall <- avg_predictions(mod_intercept, type = "response")
p_overall
```

::: {.callout-tip collapse="true"}
## Model interpretation

The estimated probability of treatment response in the full analysis sample is **`r sprintf("%.1f", 100 * p_overall$estimate[1])`%** (95% CI **`r sprintf("%.1f", 100 * p_overall$conf.low[1])`%** to **`r sprintf("%.1f", 100 * p_overall$conf.high[1])`%**).

This is a model-based estimate of an overall probability, not a treatment effect. The intercept printed by `summary()` is the log-odds corresponding to this probability.
:::

# Treatment as a binary predictor

To make a binary treatment comparison, we restrict the data to the waitlist and therapist-guided groups. Waitlist is the reference group.

```{r}
#| label: make-binary-treatment-data
df_binary <- df_clean |>
  filter(trt %in% c("waitlist", "therapist-guided")) |>
  droplevels()

df_binary |>
  count(trt, lsas_response, .drop = FALSE)
```

The logistic model is:

$$
\operatorname{logit}\left[P(Y_i = \text{Response} \mid trt_i)\right] =
\beta_0 + \beta_1\,\mathbf{1}\{trt_i = \text{therapist-guided}\}.
$$

```{r}
#| label: fit-treatment-model
mod_trt <- glm(
  lsas_response ~ trt,
  family = binomial,
  data = df_binary
)

summary(mod_trt)
```

The coefficient for therapist-guided now shows the log-odds ratio of response in the therapist-guided group vs. the waitlist group. By exponentiating, we get the odds ratio. This ban be calculated by had by dividing the odds in the therapist-guided group (31/23) by the odds in the waitlist group (6/52).

```{r}
#| label: manual-odds-ratio
exp(2.4580)
(31/23)/(6/52)
```

## Estimated probabilities,  a probability difference, and a probability ratio (a.k.a. risk ratio)

```{r}
#| label: treatment-probabilities
p_by_trt <- avg_predictions(mod_trt, by = "trt", type = "response")
p_by_trt

trt_difference <- avg_comparisons(
  mod_trt,
  variables = list(trt = c("waitlist", "therapist-guided")),
  type = "response"
)
trt_difference

trt_ratio <- avg_comparisons(
  mod_trt,
  variables = list(trt = c("waitlist", "therapist-guided")),
  type = "response",
  comparison = "ratio"
)
trt_ratio
```

::: {.callout-tip collapse="true"}
## Model interpretation

Compared with waitlist, therapist-guided treatment changes the estimated probability of treatment response by **`r sprintf("%.1f", 100 * trt_difference$estimate[1])` percentage points** (95% CI **`r sprintf("%.1f", 100 * trt_difference$conf.low[1])`** to **`r sprintf("%.1f", 100 * trt_difference$conf.high[1])` percentage points**).

The estimated probability of treatment response is **`r sprintf("%.2f", trt_ratio$estimate[1])` times as high** in the therapist-guided group as in the waitlist group (probability ratio/risk ratio; 95% CI **`r sprintf("%.2f", trt_ratio$conf.low[1])`** to **`r sprintf("%.2f", trt_ratio$conf.high[1])`**).

This is an unadjusted treatment contrast in the restricted waitlist-versus-therapist-guided sample. A positive value means that therapist-guided treatment has a higher estimated probability of treatment response.
:::

## Visualise the estimated probabilities

```{r}
#| label: plot-treatment-probabilities
plot_predictions(mod_trt, by = "trt", type = "response") +
  labs(
    title = "Estimated probability of treatment response",
    subtitle = "Binary logistic regression; 95% confidence intervals",
    x = "Treatment group",
    y = "Probability of treatment response"
  ) +
  theme_minimal()
```

# Odds, log-odds, and odds ratios

The coefficient for `trttherapist-guided` in `summary(mod_trt)` is a **log-odds difference**. Exponentiating it yields an **odds ratio**.

```{r}
#| label: treatment-odds-ratio
or_trt <- tidy(mod_trt, conf.int = TRUE, exponentiate = TRUE)
or_trt
```

An odds ratio above 1 indicates higher odds of the modelled event (`Low`) in therapist-guided treatment than waitlist. It does **not** mean that the probability is that many times larger, and it does **not** tell us the percentage-point difference.

::: callout-note
## Reading an odds ratio in a paper

Suppose an article reports an odds ratio of 2. An odds ratio of 2 means the odds are doubled—not that the probability rises by 100 percentage points, or even necessarily doubles. The corresponding probability change depends on the baseline probability. This is why we pair odds ratios with estimated probabilities and probability differences when communicating clinical results.
:::

# Summary

In this chapter you learned to:

- define a binary LSAS outcome and distinguish it from a validated clinical threshold;
- fit intercept-only and treatment logistic-regression models with `glm(..., family = binomial)`;
- interpret model output through estimated probabilities and probability differences with 95% CIs;
- visualise treatment-group probabilities with `marginaleffects`; and
- recognise an odds ratio as a ratio of odds rather than a probability difference.
