---
title: "Logistic regression with a binary and numeric predictor"
format:
  html:
    df-print: kable
toc: true
toc-depth: 3
number-sections: true
---

# A continuous moderator

In [logistic regression with two binary predictors](logistic-two-binary-predictors.qmd), baseline depression was grouped as below or at least 10 on the PHQ-9. That can be useful for a clearly defined clinical decision, but it also turns different scores into the same group. Here we retain the observed
continuous `phq9_screen` score and ask:

> How does the estimated probability of a low post-treatment LSAS score differ
> between therapist-guided treatment and waitlist at different baseline PHQ-9
> scores?

We will make that question concrete at PHQ-9 scores of 5, 10, and 15. They
represent low, moderate, and higher values within the observed range; they are
not diagnostic categories. Our primary result is a **therapist-guided minus
waitlist probability difference**, expressed in percentage points with a 95% CI.

A logistic-regression interaction is an interaction on the **log-odds scale**.
That detail matters: even an additive logistic model can have probability
differences that vary with PHQ-9, because the logit-to-probability transformation
is nonlinear. We therefore calculate contrasts on the probability scale rather
than reading their shape from a coefficient table.

# Packages and analytic data

```{r}
#| label: setup
#| message: false
#| warning: false
library(tidyverse)
library(here)
library(marginaleffects)

options(
  marginaleffects_print_omit = c("s.value"),
  marginaleffects_print_style = ifelse(interactive(), "summary", "tinytable")
)
```

```{r}
#| label: make-analysis-data
#| message: false
#| warning: false
df_binary <- read_csv2(here("data", "steps_clean.csv")) |>
  mutate(
    # The second factor level is the event modelled by binomial glm().
    lsas_post_bin = factor(
      if_else(lsas_post < 50, "Low", "High"),
      levels = c("High", "Low")
    ),
    trt = factor(
      trt,
      levels = c("waitlist", "therapist-guided", "self-guided")
    )
  ) |>
  filter(trt %in% c("waitlist", "therapist-guided")) |>
  drop_na(lsas_post_bin, phq9_screen) |>
  droplevels()

phq_values <- c(5, 10, 15)

df_binary |>
  summarise(
    participants = n(),
    minimum_phq9 = min(phq9_screen),
    maximum_phq9 = max(phq9_screen)
  )
```

`Low` means `lsas_post < 50`; `High` means `lsas_post >= 50`. This is the
Course 3 teaching outcome, not a diagnostic or remission definition. Waitlist
is the treatment reference group, and `Low` is the event whose probability is
modelled. We restrict this chapter to the waitlist and therapist-guided arms so
that every treatment contrast has one clear direction.

# Additive logistic model

## What does “additive” mean here?

The additive model has a treatment term and a PHQ-9 term, but no product term:

$$
\operatorname{logit}\{P(Y_i = \text{Low} \mid trt_i, PHQ9_i)\} =
\beta_0 +
\beta_1\mathbf{1}\{trt_i = \text{therapist-guided}\} +
\beta_2 PHQ9_i.
$$

The therapist-guided versus waitlist **log-odds** difference is the same at
every PHQ-9 score in this model. Equivalently, the treatment odds ratio is
constant across PHQ-9. That is not the same as saying the probability difference
is constant: a fixed change in log-odds maps to different probability changes
near 0, 0.5, and 1.

```{r}
#| label: fit-additive-logit
#| message: false
#| warning: false
mod_additive <- stats::glm(
  lsas_post_bin ~ trt + phq9_screen,
  family = binomial,
  data = df_binary
)

summary(mod_additive)

additive_trt_difference <- avg_comparisons(
  mod_additive,
  variables = list(trt = c("therapist-guided", "waitlist")),
  type = "response"
)
additive_trt_difference
```

::: {.callout-tip collapse="true"}
## Model interpretation

Averaged over the observed baseline PHQ-9 distribution, therapist-guided
treatment is estimated to change the probability of a low post-treatment LSAS
score by **`r sprintf("%.1f", 100 * additive_trt_difference$estimate[1])` percentage points**
compared with waitlist (95% CI **`r sprintf("%.1f", 100 * additive_trt_difference$conf.low[1])`**
to **`r sprintf("%.1f", 100 * additive_trt_difference$conf.high[1])` percentage points**).

This is therapist-guided minus waitlist; a positive value favours a higher
estimated probability of a low score with therapist guidance. It is a
population-average probability contrast over this analysis sample, conditional
on the additive model—not a coefficient interpreted as a percentage-point
change.
:::

```{r}
#| label: plot-additive-probabilities
#| fig-cap: "Additive-model estimated probabilities of a low post-treatment LSAS score, with 95% confidence intervals."
plot_predictions(
  mod_additive,
  by = c("phq9_screen", "trt"),
  type = "response"
) +
  labs(
    title = "Additive logistic model",
    subtitle = "Estimated probabilities with 95% confidence intervals",
    x = "Baseline PHQ-9 score",
    y = "Probability of low post-treatment LSAS",
    colour = "Treatment group",
    fill = "Treatment group"
  ) +
  theme_minimal()
```

The curves need not be parallel on the probability scale. The additive model
makes its parallel-lines assumption on the log-odds scale, not on this plot's
probability scale.

# Interaction model

## Let the treatment association vary with PHQ-9

The interaction model adds a product term:

$$
\begin{aligned}
\operatorname{logit}\{P(Y_i = \text{Low} \mid trt_i, PHQ9_i)\} =
&\ \beta_0 + \beta_1\mathbf{1}\{trt_i = \text{therapist-guided}\} + \\
&\ \beta_2 PHQ9_i +
\beta_3\mathbf{1}\{trt_i = \text{therapist-guided}\} \times PHQ9_i.
\end{aligned}
$$

Now the treatment log-odds difference can change with PHQ-9. The treatment
coefficient in `summary()` is the difference at PHQ-9 = 0, which is not a
useful stand-alone clinical summary here. We instead estimate probability
differences at prespecified, observed-range scores.

```{r}
#| label: fit-interaction-logit
#| message: false
#| warning: false
mod_interaction <- stats::glm(
  lsas_post_bin ~ trt * phq9_screen,
  family = binomial,
  data = df_binary
)

summary(mod_interaction)

conditional_trt_difference <- avg_comparisons(
  mod_interaction,
  variables = list(trt = c("therapist-guided", "waitlist")),
  newdata = datagrid(phq9_screen = phq_values),
  by = "phq9_screen",
  type = "response"
)
conditional_trt_difference
```

::: {.callout-tip collapse="true"}
## Model interpretation

At PHQ-9 = 5, therapist-guided minus waitlist is an estimated
**`r sprintf("%.1f", 100 * conditional_trt_difference$estimate[conditional_trt_difference$phq9_screen == 5])` percentage-point** difference in the probability of a low post-treatment LSAS score (95% CI **`r sprintf("%.1f", 100 * conditional_trt_difference$conf.low[conditional_trt_difference$phq9_screen == 5])`** to **`r sprintf("%.1f", 100 * conditional_trt_difference$conf.high[conditional_trt_difference$phq9_screen == 5])`**).

At PHQ-9 = 10, the corresponding difference is **`r sprintf("%.1f", 100 * conditional_trt_difference$estimate[conditional_trt_difference$phq9_screen == 10])` percentage points** (95% CI **`r sprintf("%.1f", 100 * conditional_trt_difference$conf.low[conditional_trt_difference$phq9_screen == 10])`** to **`r sprintf("%.1f", 100 * conditional_trt_difference$conf.high[conditional_trt_difference$phq9_screen == 10])`**); at PHQ-9 = 15 it is **`r sprintf("%.1f", 100 * conditional_trt_difference$estimate[conditional_trt_difference$phq9_screen == 15])` percentage points** (95% CI **`r sprintf("%.1f", 100 * conditional_trt_difference$conf.low[conditional_trt_difference$phq9_screen == 15])`** to **`r sprintf("%.1f", 100 * conditional_trt_difference$conf.high[conditional_trt_difference$phq9_screen == 15])`**).

Each contrast conditions on the displayed PHQ-9 value. Positive values mean a
higher modelled probability with therapist-guided treatment. The estimates and
intervals, rather than the interaction coefficient alone, describe whether the
clinically relevant probability difference changes across PHQ-9.
:::

```{r}
#| label: plot-interaction-probabilities
#| fig-cap: "Interaction-model estimated probabilities of a low post-treatment LSAS score, with 95% confidence intervals."
plot_predictions(
  mod_interaction,
  by = c("phq9_screen", "trt"),
  type = "response"
) +
  labs(
    title = "Interaction logistic model",
    subtitle = "Estimated probabilities with 95% confidence intervals",
    x = "Baseline PHQ-9 score",
    y = "Probability of low post-treatment LSAS",
    colour = "Treatment group",
    fill = "Treatment group"
  ) +
  theme_minimal()
```

## A curve of the conditional probability difference

The probability curves are useful, but they leave the reader to visually
subtract two uncertain quantities. A direct contrast plot answers the treatment
question more directly.

```{r}
#| label: plot-conditional-treatment-difference
#| fig-cap: "Conditional therapist-guided-minus-waitlist probability difference across baseline PHQ-9, with 95% confidence intervals."
plot_comparisons(
  mod_interaction,
  variables = list(trt = c("therapist-guided", "waitlist")),
  condition = "phq9_screen",
  type = "response"
) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  labs(
    title = "Conditional treatment probability difference",
    subtitle = "Therapist-guided minus waitlist; 95% confidence intervals",
    x = "Baseline PHQ-9 score",
    y = "Probability difference"
  ) +
  scale_y_continuous(labels = scales::label_percent(accuracy = 1)) +
  theme_minimal()
```

The dashed zero line marks no modelled probability difference. Wide intervals at
the edge of the observed PHQ-9 range are a reminder not to make subgroup claims
from the curve's point estimates alone.

# Conditional and population-average treatment contrasts

A conditional contrast answers a question such as “what is the modelled
difference at PHQ-9 = 10?” A population-average contrast answers a different
question: “what is the modelled treatment difference after averaging each
participant's counterfactual predictions over the observed PHQ-9 distribution?”

```{r}
#| label: interaction-marginal-treatment-difference
marginal_trt_difference <- avg_comparisons(
  mod_interaction,
  variables = list(trt = c("therapist-guided", "waitlist")),
  type = "response"
)
marginal_trt_difference
```

::: {.callout-tip collapse="true"}
## Model interpretation

Averaged over the observed PHQ-9 distribution, the interaction model estimates
that therapist-guided treatment changes the probability of a low post-treatment
LSAS score by **`r sprintf("%.1f", 100 * marginal_trt_difference$estimate[1])` percentage points**
compared with waitlist (95% CI **`r sprintf("%.1f", 100 * marginal_trt_difference$conf.low[1])`**
to **`r sprintf("%.1f", 100 * marginal_trt_difference$conf.high[1])` percentage points**).

This population-average contrast is not the effect at a particular PHQ-9 score.
It averages conditional contrasts over the PHQ-9 values represented in this
sample, so it can change if the target population's baseline PHQ-9 distribution
changes.
:::

# Odds ratios are not probability differences

The exponentiated treatment coefficient from the interaction model is an odds
ratio **at PHQ-9 = 0**. That value is outside the useful interpretive range for
this example, so reporting it alone would be particularly misleading. To see
how the odds ratio changes with PHQ-9 in the interaction model, we can estimate
the treatment log-odds difference at the same three values and exponentiate it.

```{r}
#| label: conditional-odds-ratios
conditional_trt_log_odds <- avg_comparisons(
  mod_interaction,
  variables = list(trt = c("therapist-guided", "waitlist")),
  newdata = datagrid(phq9_screen = phq_values),
  by = "phq9_screen",
  type = "link"
) |>
  transmute(
    phq9_screen,
    odds_ratio = exp(estimate),
    conf.low = exp(conf.low),
    conf.high = exp(conf.high)
  )

conditional_trt_log_odds
```

An odds ratio compares odds, $p/(1-p)$, rather than probabilities. For example,
an odds ratio of 2 means that the modelled odds are twice as large; it does not
mean that the probability rises by 100 percentage points or necessarily doubles.
The probability change depends on the starting probability. For the clinical
question in this chapter, report the conditional probability differences and
their CIs first; use the odds ratios to read and critique the statistical
literature accurately.

# Summary

- An additive logistic model assumes a common treatment **log-odds** difference,
  not necessarily a constant probability difference.
- An interaction model lets the treatment association vary with continuous
  baseline PHQ-9.
- `avg_comparisons(..., type = "response")` supplies conditional and
  population-average treatment contrasts on the probability scale.
- A contrast curve communicates both the estimated pattern and its uncertainty.
- Odds ratios describe ratios of odds. They are not percentage-point changes in
  probability.
