---
title: "Logistic-regression foundations"
format:
  html:
    df-print: kable
toc: true
toc-depth: 3
number-sections: true
---

# Why logistic regression?

Course 2 used regression to model continuous outcomes such as post-treatment LSAS scores. In many psychiatric research questions, the outcome is instead binary:

- remission versus non-remission;
- response versus non-response;
- attendance versus dropout; or
- an adverse event versus no adverse event.

A linear model can produce values below 0 or above 1 when applied to a binary outcome. Logistic regression instead models a probability while keeping model-based predictions between 0 and 1.

This chapter uses a binary post-treatment social-anxiety outcome. Our main question is:

> How does the probability of a low post-treatment LSAS score differ between treatment groups?

The answer will be expressed primarily as estimated probabilities and **probability differences in percentage points**. We will also inspect odds ratios, because they are common in the literature, but an odds ratio is not a probability difference.

# Packages and data

```{r}
#| label: setup
#| message: false
#| warning: false
library(tidyverse)
library(here)
library(broom)
library(marginaleffects)

options(
  marginaleffects_print_omit = c("s.value"),
  marginaleffects_print_style = ifelse(interactive(), "summary", "tinytable")
)
```

```{r}
#| label: load-data
#| message: false
#| warning: false
df_clean <- read_csv2(here("data", "steps_clean.csv")) |>
  mutate(
    # The second factor level is the event modelled by binomial glm().
    lsas_post_bin = factor(
      if_else(lsas_post < 50, "Low", "High"),
      levels = c("High", "Low")
    ),
    trt = factor(
      trt,
      levels = c("waitlist", "therapist-guided", "self-guided")
    )
  )

df_clean |>
  count(lsas_post_bin, .drop = FALSE)
```

The binary outcome is deliberately simple:

```r
lsas_post_bin = "Low"  if lsas_post < 50
lsas_post_bin = "High" if lsas_post >= 50
```

This matches the Course 1 classification example. It is useful for learning, but it discards information: LSAS scores of 49 and 12 are both labelled `Low`, while scores of 50 and 120 are both labelled `High`. The original continuous score remains important.

# Thresholds, categories, and humility

The LSAS-SR is often used to screen for social anxiety disorder. The original clinical LSAS study found total-score cut-points of 30 for social anxiety disorder and 60 for the then-defined generalised subtype [@Mennin2002LSAS]. A study of the self-report LSAS-SR found the same cut-points had a useful sensitivity/specificity balance [@Rytwinski2009LSASSR].

Thresholds can also serve different purposes. In a treated German social-anxiety sample, an LSAS-SR score of 35 was an empirically derived remission cut-point, and a 28% reduction identified response [@vonGlischinski2018LSAS]. Those are not automatically the right definitions for every study or clinical decision.

::: {.callout-caution}
## A threshold is an analysis decision

The `<50` teaching outcome is neither a diagnosis nor a remission definition. Before dichotomising a clinical scale in your own work, specify the purpose, identify a defensible external threshold where one exists, and report the information lost by categorisation.
:::

# Intercept-only logistic regression

We first ignore treatment group and estimate one probability: the probability that a participant has a low LSAS score at post-treatment.

The model is:

$$
\operatorname{logit}\left[P(Y_i = \text{Low})\right] = \beta_0,
$$

where the logit transforms a probability $p$ into log-odds:

$$
\operatorname{logit}(p) = \log\left(\frac{p}{1-p}\right).
$$

```{r}
#| label: fit-intercept-only
mod_intercept <- glm(
  lsas_post_bin ~ 1,
  family = binomial,
  data = df_clean
)

summary(mod_intercept)
```

The intercept is on the log-odds scale, so it is not the clearest clinical summary. We use `marginaleffects` to obtain the estimated probability and its confidence interval.

```{r}
#| label: intercept-probability
p_overall <- avg_predictions(mod_intercept, type = "response")
p_overall
```

::: {.callout-tip collapse="true"}
## Model interpretation

The estimated probability of a low post-treatment LSAS score in the full analysis sample is **`r sprintf("%.1f", 100 * p_overall$estimate[1])`%** (95% CI **`r sprintf("%.1f", 100 * p_overall$conf.low[1])`%** to **`r sprintf("%.1f", 100 * p_overall$conf.high[1])`%**).

This is a model-based estimate of an overall probability, not a treatment effect. The intercept printed by `summary()` is the log-odds corresponding to this probability.
:::

# Treatment as a binary predictor

To make a binary treatment comparison, we restrict the data to the waitlist and therapist-guided groups. Waitlist is the reference group.

```{r}
#| label: make-binary-treatment-data
df_binary <- df_clean |>
  filter(trt %in% c("waitlist", "therapist-guided")) |>
  droplevels()

df_binary |>
  count(trt, lsas_post_bin, .drop = FALSE)
```

The logistic model is:

$$
\operatorname{logit}\left[P(Y_i = \text{Low} \mid trt_i)\right] =
\beta_0 + \beta_1\,\mathbf{1}\{trt_i = \text{therapist-guided}\}.
$$

```{r}
#| label: fit-treatment-model
mod_trt <- glm(
  lsas_post_bin ~ trt,
  family = binomial,
  data = df_binary
)

summary(mod_trt)
```

## Estimated probabilities and a probability difference

```{r}
#| label: treatment-probabilities
p_by_trt <- avg_predictions(mod_trt, by = "trt", type = "response")
p_by_trt

trt_difference <- avg_comparisons(
  mod_trt,
  variables = list(trt = c("therapist-guided", "waitlist")),
  type = "response"
)
trt_difference
```

::: {.callout-tip collapse="true"}
## Model interpretation

Compared with waitlist, therapist-guided treatment changes the estimated probability of a low post-treatment LSAS score by **`r sprintf("%.1f", 100 * trt_difference$estimate[1])` percentage points** (95% CI **`r sprintf("%.1f", 100 * trt_difference$conf.low[1])`** to **`r sprintf("%.1f", 100 * trt_difference$conf.high[1])` percentage points**).

This is an unadjusted treatment contrast in the restricted waitlist-versus-therapist-guided sample. A positive value means that therapist-guided treatment has a higher estimated probability of a low post-treatment LSAS score.
:::

## Visualise the estimated probabilities

```{r}
#| label: plot-treatment-probabilities
plot_predictions(mod_trt, by = "trt", type = "response") +
  labs(
    title = "Estimated probability of low post-treatment LSAS",
    subtitle = "Binary logistic regression; 95% confidence intervals",
    x = "Treatment group",
    y = "Probability of low post-treatment LSAS"
  ) +
  theme_minimal()
```

# Odds, log-odds, and odds ratios

The coefficient for `trttherapist-guided` in `summary(mod_trt)` is a **log-odds difference**. Exponentiating it yields an **odds ratio**.

```{r}
#| label: treatment-odds-ratio
or_trt <- tidy(mod_trt, conf.int = TRUE, exponentiate = TRUE)
or_trt
```

An odds ratio above 1 indicates higher odds of the modelled event (`Low`) in therapist-guided treatment than waitlist. It does **not** mean that the probability is that many times larger, and it does **not** tell us the percentage-point difference.

::: {.callout-note}
## Reading an odds ratio in a paper

Suppose an article reports an odds ratio of 2. An odds ratio of 2 means the odds are doubled—not that the probability rises by 100 percentage points, or even necessarily doubles. The corresponding probability change depends on the baseline probability. This is why we pair odds ratios with estimated probabilities and probability differences when communicating clinical results.
:::

# Summary

In this chapter you learned to:

- define a binary LSAS outcome and distinguish it from a validated clinical threshold;
- fit intercept-only and treatment logistic-regression models with `glm(..., family = binomial)`;
- interpret model output through estimated probabilities and probability differences with 95% CIs;
- visualise treatment-group probabilities with `marginaleffects`; and
- recognise an odds ratio as a ratio of odds rather than a probability difference.
