---
title: "Logistic regression with a binary predictor"
format:
  html:
    df-print: kable
toc: true
toc-depth: 3
number-sections: true
---

# Introduction

This notebook extends the treatment example from
[logistic-regression foundations](logistic-intro.qmd). We focus on the two arms
that make a binary treatment predictor: `waitlist` and `therapist-guided`.

Our question is:

> Among participants in these two arms, how does the estimated probability of
> a low post-treatment LSAS score differ for therapist-guided treatment versus
> waitlist?

The target is the **therapist-guided minus waitlist probability difference**.
We will use the corresponding odds ratio to read logistic-regression output,
but the probability difference is the primary clinical result.

# Packages and data

```{r}
#| label: setup
#| message: false
#| warning: false
library(tidyverse)
library(here)
library(broom)
library(marginaleffects)

options(
  marginaleffects_print_omit = c("s.value"),
  marginaleffects_print_style = ifelse(interactive(), "summary", "tinytable")
)
```

```{r}
#| label: load-binary-treatment-data
#| message: false
#| warning: false
df_binary <- read_csv2(here("data", "steps_clean.csv")) |>
  mutate(
    # The second level is the event modelled by binomial glm().
    lsas_post_bin = factor(
      if_else(lsas_post < 50, "Low", "High"),
      levels = c("High", "Low")
    ),
    trt = factor(
      trt,
      levels = c("waitlist", "therapist-guided", "self-guided")
    )
  ) |>
  filter(trt %in% c("waitlist", "therapist-guided")) |>
  droplevels()

df_binary |>
  count(trt, lsas_post_bin, .drop = FALSE)
```

`Low` means `lsas_post < 50`; `High` means `lsas_post >= 50`. As in
[logistic-regression foundations](logistic-intro.qmd),
this is a teaching outcome, not a diagnosis or a remission definition. The
factor order makes `High` the first level, so that the binomial model estimates
the probability of the second level: a low post-treatment LSAS score.

We explicitly make waitlist the reference group. The treatment contrast shown
by the coefficient will therefore be therapist-guided minus waitlist.

```{r}
#| label: inspect-binary-treatment-coding
levels(df_binary$trt)
contrasts(df_binary$trt)
```

# A logistic model for two treatment groups

For participant $i$, let $Y_i$ indicate a low post-treatment LSAS score. The
model is

$$
\operatorname{logit}\left[P(Y_i = \text{Low} \mid trt_i)\right] =
\beta_0 + \beta_1\,\mathbf{1}\{trt_i = \text{therapist-guided}\}.
$$

Here, $\beta_0$ is the log-odds of a low score in the waitlist group, and
$\beta_1$ is the therapist-guided-versus-waitlist log-odds difference. These
are the quantities that `glm()` estimates directly. They are not yet the most
clinically intelligible quantities.

| Treatment group | Linear predictor | Model-derived probability |
|---|---:|---:|
| waitlist | $\beta_0$ | $\operatorname{logit}^{-1}(\beta_0)$ |
| therapist-guided | $\beta_0 + \beta_1$ | $\operatorname{logit}^{-1}(\beta_0 + \beta_1)$ |

## Fit and interpret the model

```{r}
#| label: fit-binary-treatment-logit
#| message: false
#| warning: false
mod_binary_trt <- stats::glm(
  lsas_post_bin ~ trt,
  family = binomial,
  data = df_binary
)

p_binary_trt <- avg_predictions(
  mod_binary_trt,
  by = "trt",
  type = "response"
)

diff_binary_trt <- avg_comparisons(
  mod_binary_trt,
  variables = list(trt = c("waitlist", "therapist-guided")),
  type = "response"
)

or_binary_trt <- tidy(
  mod_binary_trt,
  conf.int = TRUE,
  exponentiate = TRUE
) |>
  filter(term == "trttherapist-guided")
```

::: {.callout-tip collapse="true"}
## Model interpretation

The estimated probability of a low post-treatment LSAS score is
**`r sprintf("%.1f", 100 * p_binary_trt$estimate[p_binary_trt$trt == "waitlist"])`%**
for waitlist (95% CI **`r sprintf("%.1f", 100 * p_binary_trt$conf.low[p_binary_trt$trt == "waitlist"])`%**
to **`r sprintf("%.1f", 100 * p_binary_trt$conf.high[p_binary_trt$trt == "waitlist"])`%**) and
**`r sprintf("%.1f", 100 * p_binary_trt$estimate[p_binary_trt$trt == "therapist-guided"])`%**
for therapist-guided treatment (95% CI **`r sprintf("%.1f", 100 * p_binary_trt$conf.low[p_binary_trt$trt == "therapist-guided"])`%**
to **`r sprintf("%.1f", 100 * p_binary_trt$conf.high[p_binary_trt$trt == "therapist-guided"])`%**).

Compared with waitlist, therapist-guided treatment **`r ifelse(diff_binary_trt$estimate[1] >= 0, "increases", "decreases")`**
the estimated probability by **`r sprintf("%.1f", abs(100 * diff_binary_trt$estimate[1]))` percentage points**
(therapist-guided minus waitlist = **`r sprintf("%.1f", 100 * diff_binary_trt$estimate[1])` percentage points**;
95% CI **`r sprintf("%.1f", 100 * diff_binary_trt$conf.low[1])`** to
**`r sprintf("%.1f", 100 * diff_binary_trt$conf.high[1])` percentage points**).
This is an unadjusted contrast within the restricted two-arm sample.

The associated therapist-guided-versus-waitlist **odds ratio** is
**`r sprintf("%.2f", or_binary_trt$estimate[1])`** (95% CI
**`r sprintf("%.2f", or_binary_trt$conf.low[1])`** to
**`r sprintf("%.2f", or_binary_trt$conf.high[1])`**). It is a ratio of the
odds of a low score, not a probability ratio and not a
`r sprintf("%.1f", 100 * diff_binary_trt$estimate[1])`-percentage-point change.
:::

## Display probabilities and the probability difference

```{r}
#| label: show-binary-treatment-estimates
p_binary_trt

diff_binary_trt
```

The estimated probability difference is calculated after transforming the
model to the response scale. It answers a different question from the printed
log-odds coefficient: how many percentage points higher or lower is the
modelled probability in one group than the other?

## Visualise the group probabilities

```{r}
#| label: plot-binary-treatment-probabilities
plot_predictions(mod_binary_trt, by = "trt", type = "response") +
  labs(
    title = "Estimated probability of low post-treatment LSAS",
    subtitle = "Waitlist versus therapist-guided treatment; 95% confidence intervals",
    x = "Treatment group",
    y = "Probability of low post-treatment LSAS"
  ) +
  theme_minimal()
```

# Log-odds and odds ratios

The logistic model is linear on the **log-odds** scale. Exponentiating
$\beta_1$ gives the therapist-guided-versus-waitlist odds ratio:

$$
\exp(\beta_1) =
\frac{\text{odds of a low score with therapist-guided treatment}}
{\text{odds of a low score with waitlist}}.
$$

```{r}
#| label: show-binary-treatment-odds-ratio
or_binary_trt
```

An odds ratio larger than 1 means higher *odds* of a low score in the
therapist-guided group. It cannot be read as a percentage-point increase,
because the same odds ratio implies different probability differences at
different baseline probabilities. Report the model-derived probabilities and
their probability difference when the clinical question concerns chances of an
outcome.

# Summary

In this chapter you learned to:

- restrict the STePS data to a binary treatment comparison and set waitlist as
the reference category;
- fit a binary logistic model with `stats::glm(..., family = binomial)`;
- report treatment-specific probabilities and the therapist-guided-minus-waitlist
probability difference with 95% CIs; and
- interpret the associated odds ratio as a ratio of odds rather than a
probability difference.
