---
title: "Logistic regression with a categorical predictor"
format:
  html:
    df-print: kable
toc: true
toc-depth: 3
number-sections: true
---

# Introduction

A treatment variable with three arms is a **categorical predictor**, not a
single binary treatment indicator. In the STePS data, the arms are waitlist,
therapist-guided treatment, and self-guided treatment.

Our primary questions are:

1. What is the model-derived probability of treatment response in
   each treatment arm?
2. How do those probabilities differ between clinically relevant pairs of arms?

As before, we will communicate probabilities and probability differences with
95% CIs. Log-odds coefficients and odds ratios help us read model output, but
they are not probability differences.

# Packages and data

```{r}
#| label: setup
#| message: false
#| warning: false
library(tidyverse)
library(here)
library(broom)
library(marginaleffects)

options(
  marginaleffects_print_omit = c("s.value"),
  marginaleffects_print_style = ifelse(interactive(), "summary", "tinytable")
)
```

```{r}
#| label: load-categorical-treatment-data
#| message: false
#| warning: false
df_categorical <- read_csv2(here("data", "steps_clean.csv")) |>
  filter(!is.na(lsas_screen), !is.na(lsas_post)) |>
  mutate(
    # The second level is the event modelled by binomial glm().
    lsas_response = factor(
      if_else((lsas_screen - lsas_post) / lsas_screen >= 0.30, "Response", "No response"),
      levels = c("No response", "Response")
    ),
    trt = factor(
      trt,
      levels = c("waitlist", "therapist-guided", "self-guided")
    )
  )

df_categorical |>
  count(trt, lsas_response, .drop = FALSE)
```

`Response` is a reduction of at least 30% from baseline LSAS; it is not a diagnostic or remission definition. We order
the outcome so that `Low` is the event in the binomial model, and we explicitly
set `waitlist` as the treatment reference category.

```{r}
#| label: inspect-categorical-treatment-coding
levels(df_categorical$trt)
contrasts(df_categorical$trt)
```

# Reference categories and dummy variables

With waitlist as the reference, R uses two indicator variables. The model is

$$
\operatorname{logit}\left[P(Y_i = \text{Response} \mid trt_i)\right] =
\beta_0 +
\beta_1\,\mathbf{1}\{trt_i = \text{therapist-guided}\} +
\beta_2\,\mathbf{1}\{trt_i = \text{self-guided}\}.
$$

| Treatment group | Linear predictor | Coefficient comparison |
|---|---:|---|
| waitlist | $\beta_0$ | reference |
| therapist-guided | $\beta_0 + \beta_1$ | therapist-guided minus waitlist |
| self-guided | $\beta_0 + \beta_2$ | self-guided minus waitlist |

Thus, the coefficient table supplies two log-odds comparisons with waitlist.
It does **not** directly show the self-guided-versus-therapist-guided
comparison; we obtain that, and all other pairwise probability differences,
from the fitted model.

# Fit the three-arm logistic model

```{r}
#| label: fit-categorical-treatment-logit
#| message: false
#| warning: false
mod_categorical_trt <- stats::glm(
  lsas_response ~ trt,
  family = binomial,
  data = df_categorical
)

p_categorical_trt <- avg_predictions(
  mod_categorical_trt,
  by = "trt",
  type = "response"
)

diff_guided_waitlist <- avg_comparisons(
  mod_categorical_trt,
  variables = list(trt = c("waitlist", "therapist-guided")),
  type = "response"
)

diff_self_waitlist <- avg_comparisons(
  mod_categorical_trt,
  variables = list(trt = c("waitlist", "self-guided")),
  type = "response"
)

or_categorical_trt <- tidy(
  mod_categorical_trt,
  conf.int = TRUE,
  exponentiate = TRUE
) |>
  filter(term != "(Intercept)")
```

::: {.callout-tip collapse="true"}
## Model interpretation

The estimated probability of treatment response is
**`r sprintf("%.1f", 100 * p_categorical_trt$estimate[p_categorical_trt$trt == "waitlist"])`%**
for waitlist (95% CI **`r sprintf("%.1f", 100 * p_categorical_trt$conf.low[p_categorical_trt$trt == "waitlist"])`%**
to **`r sprintf("%.1f", 100 * p_categorical_trt$conf.high[p_categorical_trt$trt == "waitlist"])`%**),
**`r sprintf("%.1f", 100 * p_categorical_trt$estimate[p_categorical_trt$trt == "therapist-guided"])`%**
for therapist-guided treatment (95% CI **`r sprintf("%.1f", 100 * p_categorical_trt$conf.low[p_categorical_trt$trt == "therapist-guided"])`%**
to **`r sprintf("%.1f", 100 * p_categorical_trt$conf.high[p_categorical_trt$trt == "therapist-guided"])`%**), and
**`r sprintf("%.1f", 100 * p_categorical_trt$estimate[p_categorical_trt$trt == "self-guided"])`%**
for self-guided treatment (95% CI **`r sprintf("%.1f", 100 * p_categorical_trt$conf.low[p_categorical_trt$trt == "self-guided"])`%**
to **`r sprintf("%.1f", 100 * p_categorical_trt$conf.high[p_categorical_trt$trt == "self-guided"])`%**).

Relative to waitlist, therapist-guided treatment **`r ifelse(diff_guided_waitlist$estimate[1] >= 0, "increases", "decreases")`**
the estimated probability by **`r sprintf("%.1f", abs(100 * diff_guided_waitlist$estimate[1]))` percentage points**
(95% CI **`r sprintf("%.1f", 100 * diff_guided_waitlist$conf.low[1])`** to
**`r sprintf("%.1f", 100 * diff_guided_waitlist$conf.high[1])` percentage points**).
Self-guided treatment **`r ifelse(diff_self_waitlist$estimate[1] >= 0, "increases", "decreases")`**
it by **`r sprintf("%.1f", abs(100 * diff_self_waitlist$estimate[1]))` percentage points**
relative to waitlist (95% CI **`r sprintf("%.1f", 100 * diff_self_waitlist$conf.low[1])`** to
**`r sprintf("%.1f", 100 * diff_self_waitlist$conf.high[1])` percentage points**).
These are unadjusted contrasts in the three-arm analysis sample.

The corresponding odds ratios versus waitlist are
**`r sprintf("%.2f", or_categorical_trt$estimate[or_categorical_trt$term == "trttherapist-guided"])`**
for therapist-guided treatment (95% CI
**`r sprintf("%.2f", or_categorical_trt$conf.low[or_categorical_trt$term == "trttherapist-guided"])`** to
**`r sprintf("%.2f", or_categorical_trt$conf.high[or_categorical_trt$term == "trttherapist-guided"])`**) and
**`r sprintf("%.2f", or_categorical_trt$estimate[or_categorical_trt$term == "trtself-guided"])`**
for self-guided treatment (95% CI
**`r sprintf("%.2f", or_categorical_trt$conf.low[or_categorical_trt$term == "trtself-guided"])`** to
**`r sprintf("%.2f", or_categorical_trt$conf.high[or_categorical_trt$term == "trtself-guided"])`**).
Each is a ratio of odds, not a probability ratio or a percentage-point
probability difference.
:::

# Predicted probabilities by treatment arm

```{r}
#| label: show-categorical-treatment-probabilities
p_categorical_trt
```

The three predicted probabilities are response-scale summaries of the same
logistic model. They are often a more direct starting point for a clinical
discussion than its intercept and two log-odds coefficients.

```{r}
#| label: plot-categorical-treatment-probabilities
plot_predictions(mod_categorical_trt, by = "trt", type = "response") +
  labs(
    title = "Estimated probability of treatment response",
    subtitle = "Three-arm logistic regression; 95% confidence intervals",
    x = "Treatment group",
    y = "Probability of treatment response"
  ) +
  theme_minimal()
```

# Probability differences between treatment arms

There is no single treatment contrast with three arms. Decide which comparison
answers the substantive question, state its direction, and present uncertainty.
For an exploratory overview, calculate all pairwise differences. A positive
difference means that the first-named group has the higher estimated probability
of treatment response.

```{r}
#| label: pairwise-categorical-probability-differences
pairwise_categorical_differences <- avg_comparisons(
  mod_categorical_trt,
  variables = list(trt = "pairwise"),
  type = "response"
)

pairwise_categorical_differences
```

```{r}
#| label: plot-pairwise-categorical-probability-differences
pairwise_categorical_differences |>
  mutate(
    estimate_pp = 100 * estimate,
    conf.low_pp = 100 * conf.low,
    conf.high_pp = 100 * conf.high
  ) |>
  ggplot(aes(y = contrast, x = estimate_pp)) +
  geom_vline(xintercept = 0, linetype = "dashed") +
  geom_errorbar(aes(xmin = conf.low_pp, xmax = conf.high_pp), width = 0.1) +
  geom_point(size = 2.5) +
  labs(
    title = "Pairwise differences in probability of treatment response",
    subtitle = "First-named treatment minus second-named treatment; 95% confidence intervals",
    x = "Probability difference (percentage points)",
    y = NULL
  ) +
  theme_minimal()
```

# Odds ratios as a secondary summary

```{r}
#| label: show-categorical-treatment-odds-ratios
or_categorical_trt
```

The odds ratios above compare each active arm with waitlist because waitlist is
the reference group. To obtain a self-guided-versus-therapist-guided odds
ratio, one could change the reference category or form the appropriate model
contrast. Neither route changes the main communication task: accompany any odds
ratio with response-scale probabilities and a clearly directed probability
difference.

# Summary

In this chapter you learned to:

- define an explicit reference category for a three-level treatment predictor;
- fit a three-arm binary logistic model with `stats::glm(..., family = binomial)`;
- obtain estimated probabilities for every treatment arm and pairwise
probability differences with 95% CIs; and
- distinguish odds ratios from the percentage-point probability differences
needed for a clinical interpretation.
