---
title: "Logistic regression with two binary predictors"
format:
  html:
    df-print: kable
toc: true
toc-depth: 3
number-sections: true
---

# Two predictors, one clinical question

The previous logistic-regression chapters considered one predictor at a time. Clinical questions often ask whether a treatment contrast looks different in a baseline subgroup. Here we model the probability of a **treatment response** using:

- treatment group (`trt`); and
- a binary baseline PHQ-9 category (`phq_cat`).

Our teaching question is:

> Within each PHQ-9 category, how does the estimated probability of treatment response differ between therapist-guided treatment and waitlist?

We will first fit an additive logistic model and then a model with `trt * phq_cat`. The probability difference, with its 95% confidence interval, is the clinical contrast. Log-odds and odds ratios help us read model output and research papers, but neither is a percentage-point difference.

# Packages and data

```{r}
#| label: setup
#| message: false
#| warning: false
library(tidyverse)
library(here)
library(broom)
library(marginaleffects)

options(
  marginaleffects_print_omit = c("s.value"),
  marginaleffects_print_style = ifelse(interactive(), "summary", "tinytable")
)
```

```{r}
#| label: load-and-derive-data
#| message: false
#| warning: false
df_binary <- read_csv2(here("data", "steps_clean.csv")) |>
  filter(!is.na(lsas_screen), !is.na(lsas_post)) |>
  mutate(
    # The second outcome level is the event modelled by binomial glm().
    lsas_response = factor(
      if_else(
        (lsas_screen - lsas_post) / lsas_screen >= 0.30,
        "Response",
        "No response"
      ),
      levels = c("No response", "Response")
    ),
    trt = factor(
      trt,
      levels = c("waitlist", "therapist-guided", "self-guided")
    ),
    phq_cat = factor(
      case_when(
        phq9_screen < 10 ~ "< 10",
        phq9_screen >= 10 ~ ">= 10"
      ),
      levels = c("< 10", ">= 10")
    )
  ) |>
  filter(trt %in% c("waitlist", "therapist-guided")) |>
  droplevels()

df_binary |>
  count(phq_cat, lsas_response, .drop = FALSE)
```

`phq_cat` explicitly divides baseline PHQ-9 scores into `< 10` and `>= 10`. It is a convenient binary moderator for this lesson, not a claim that these two groups are biologically distinct or that a subgroup result is clinically established.
Splitting PHQ-9 at 10 also discards information: a participant scoring 10 is grouped with one scoring 25, and apart from one scoring 9 [@Altman2006Dichotomising; @MacCallum2002Dichotomization].
[Logistic regression with a binary and numeric predictor](logistic-binary-numeric-predictors.qmd) keeps PHQ-9 continuous.
The outcome is treatment response, defined as a reduction of at least 30% from baseline LSAS.

Waitlist is the reference treatment group, `< 10` is the reference PHQ-9 category, and `Response` is the modelled outcome event.

# Additive logistic model

An additive model has treatment and PHQ-9-category terms but no product term:

$$
\operatorname{logit}\left[P(Y_i = \mathrm{Response})\right]
=
\beta_0 + \beta_1 trt_i + \beta_2 phq_i
$$

On the **log-odds scale**, this model says that the treatment odds ratio is the same in both PHQ-9 categories.

```{r}
#| label: fit-additive-logit
#| message: false
#| warning: false
mod_additive <- stats::glm(
  lsas_response ~ trt + phq_cat,
  family = binomial,
  data = df_binary
)

summary(mod_additive)
```

We can get the OR of both therapist-guided vs waitlist and PHQ-9 high vs low by exponentiating the coefficients. Because the additive model does not allow for interactions on the log-odds or odds scale, the OR between the treatment groups are forced to be the same, regardless of PHQ-9 score.

```{r}
#| label: fit-additive-logit-manual
b_additive <- coef(mod_additive)
exp(b_additive[["trttherapist-guided"]]) # OR therapist-guided vs waitlist
exp(b_additive[["phq_cat>= 10"]]) # OR PHQ-9 >= 10 vs < 10
```

The fact that the odds ratio is the same in both PHQ-9 categories does *not* guarantee identical probability differences: changing a probability by a fixed odds ratio can produce different percentage-point changes at different baseline probabilities.

```{r}
#| label: fit-additive-logit-probabilities
#| message: false
#| warning: false

additive_probabilities <- avg_predictions(
  mod_additive,
  by = c("trt", "phq_cat"),
  type = "response"
)

additive_treatment_effects <- avg_comparisons(
  mod_additive,
  variables = list(trt = c("waitlist", "therapist-guided")),
  by = "phq_cat",
  type = "response"
)

additive_treatment_low_phq <- additive_treatment_effects |>
  filter(phq_cat == "< 10")
additive_treatment_high_phq <- additive_treatment_effects |>
  filter(phq_cat == ">= 10")

additive_odds_ratio <- tidy(
  mod_additive,
  conf.int = TRUE,
  exponentiate = TRUE
) |>
  filter(term == "trttherapist-guided")

additive_probabilities
additive_treatment_effects
```

::: {.callout-tip collapse="true"}
## Model interpretation

Compared with waitlist, therapist-guided treatment has an estimated probability of treatment response that is **`r if (additive_treatment_low_phq$estimate[1] >= 0) "higher" else "lower"` by `r sprintf("%.1f", 100 * abs(additive_treatment_low_phq$estimate[1]))` percentage points** among participants with baseline PHQ-9 `< 10` (95% CI **`r sprintf("%.1f", 100 * additive_treatment_low_phq$conf.low[1])` to `r sprintf("%.1f", 100 * additive_treatment_low_phq$conf.high[1])` percentage points**). Among participants with baseline PHQ-9 `>= 10`, it is **`r if (additive_treatment_high_phq$estimate[1] >= 0) "higher" else "lower"` by `r sprintf("%.1f", 100 * abs(additive_treatment_high_phq$estimate[1]))` percentage points** (95% CI **`r sprintf("%.1f", 100 * additive_treatment_high_phq$conf.low[1])` to `r sprintf("%.1f", 100 * additive_treatment_high_phq$conf.high[1])` percentage points**).

Because this is an additive *logistic* model, its therapist-guided-versus-waitlist odds ratio is **`r sprintf("%.2f", additive_odds_ratio$estimate[1])`** (95% CI **`r sprintf("%.2f", additive_odds_ratio$conf.low[1])` to `r sprintf("%.2f", additive_odds_ratio$conf.high[1])`**) in both PHQ-9 categories, adjusted for PHQ-9 category. This is a ratio of odds, not a percentage-point probability change; the conditional probability differences above are the primary results.
:::

```{r}
#| label: plot-additive-probabilities
#| fig-cap: "Additive-model estimated probabilities of treatment response. Error bars are 95% confidence intervals."
plot_predictions(
  mod_additive,
  by = c("trt", "phq_cat"),
  type = "response"
) +
  labs(
    title = "Additive logistic model",
    subtitle = "Estimated probabilities by treatment and baseline PHQ-9 category",
    x = "Treatment group",
    y = "Estimated probability of treatment response (95% CI)",
    colour = "Baseline PHQ-9"
  ) +
  theme_minimal()
```

# Interaction model

::: callout-note
## Effect modification, interaction, and moderation

These terms are often used interchangeably, but they mean different things.

- **Effect modification**: the effect of one intervention (here, treatment) differs across levels of another variable (here, baseline PHQ-9 category). Only the first variable needs to be something we could intervene on; the second is a characteristic we condition on [@VanderWeele2009Interaction].
- **Causal interaction**: the effect of intervening on *both* variables together differs from what their separate effects would imply. Both variables must be things we could, at least in principle, intervene on [@VanderWeele2009Interaction].
- **Moderation**: the term most used in psychology for a variable that changes the direction or strength of a relationship [@Baron1986Moderator]. In randomized trials, @Kraemer2002Moderators define a moderator as a characteristic measured before randomization, and therefore unrelated to treatment, that identifies subgroups with different treatment effects. This is close to effect modification.

Treatment is randomized, but PHQ-9 is not, so this chapter studies effect modification.
We can estimate whether the treatment effect differs between PHQ-9 categories, but not what would happen if we changed a participant's depression, and a difference between the categories may reflect other characteristics that go with higher PHQ-9 scores.

All three depend on the scale.
A model can show no interaction on one scale and an interaction on another [@VanderWeele2014Tutorial].
The additive model above is an example: it has no interaction on the log-odds scale, yet its probability differences differ between the PHQ-9 categories.
When you report effect modification, name the scale, for example the probability difference (an additive scale) or the odds ratio (a multiplicative scale).
:::

The interaction model uses the requested product term, `trt * phq_cat`:

$$
\operatorname{logit}\left[P(Y_i = \mathrm{Response})\right]
=
\beta_0 + \beta_1 trt_i + \beta_2 phq_i
+ \beta_3(trt_i \times phq_i).
$$

The interaction terms let treatment associations with the outcome differ by baseline PHQ-9 category. Since both trt and phq_cat are binary (0/1), the only combination that does not become 0 is when both are 1 (i.e. when trt = therapist-guided group and phq_cat = \>= 10). The treatment main-effect coefficients are now interpreted at the reference category, PHQ-9 `< 10`; they are not an overall treatment effect.

```{r}
#| label: fit-interaction-logit
#| message: false
#| warning: false
mod_interaction <- stats::glm(
  lsas_response ~ trt * phq_cat,
  family = binomial,
  data = df_binary
)

summary(mod_interaction)
```

We can exponentiate the coefficients to express them on the odds-ratio scale. In an interaction model, however, the treatment coefficient now represents the treatment effect when PHQ-9 is in its reference category (\< 10). Thus, $\exp(\beta_1)$ is the OR for therapist-guided treatment versus waitlist among participants with PHQ-9 \< 10.

The interaction coefficient tells us how the treatment OR changes when moving from PHQ-9 \< 10 to PHQ-9 ≥ 10. Here, $\exp(\beta_3)$ is the ratio of the two treatment ORs. To obtain the treatment OR among participants with PHQ-9 ≥ 10, we therefore combine the treatment and interaction coefficients: $\exp(\beta_1 + \beta_3)$.

```{r}
#| label: fit-interaction-logit-manual
b_interaction <- coef(mod_interaction)
b1 <- b_interaction[["trttherapist-guided"]]
b3 <- b_interaction[["trttherapist-guided:phq_cat>= 10"]]

exp(b1) # treatment OR when PHQ-9 < 10
exp(b3) # ratio of treatment ORs
exp(b1 + b3) # treatment OR when PHQ-9 >= 10
```

To go from the odds scale to the probability scale, we can again use functions from the `marginaleffects` package:

```{r}
#| label: fit-interaction-logit-probabilities
#| message: false
#| warning: false
interaction_probabilities <- avg_predictions(
  mod_interaction,
  by = c("trt", "phq_cat"),
  type = "response"
)

interaction_probabilities

interaction_treatment_effects <- avg_comparisons(
  mod_interaction,
  variables = list(trt = c("waitlist", "therapist-guided")),
  by = "phq_cat",
  type = "response"
)

interaction_treatment_effects
```

```{r}
#| label: interaction-inline-values
#| echo: false
#| output: false

interaction_treatment_low_phq <- interaction_treatment_effects |>
  filter(phq_cat == "< 10")
interaction_treatment_high_phq <- interaction_treatment_effects |>
  filter(phq_cat == ">= 10")

interaction_log_odds_effects <- avg_comparisons(
  mod_interaction,
  variables = list(trt = c("waitlist", "therapist-guided")),
  by = "phq_cat",
  type = "link"
) |>
  mutate(
    odds_ratio = exp(estimate),
    odds_ratio_low = exp(conf.low),
    odds_ratio_high = exp(conf.high)
  )

interaction_odds_ratio_low_phq <- interaction_log_odds_effects |>
  filter(phq_cat == "< 10")
interaction_odds_ratio_high_phq <- interaction_log_odds_effects |>
  filter(phq_cat == ">= 10")
```

::: {.callout-tip collapse="true"}
## Model interpretation

Within the baseline PHQ-9 `< 10` category, therapist-guided treatment has an estimated probability of treatment response that is **`r if (interaction_treatment_low_phq$estimate[1] >= 0) "higher" else "lower"` by `r sprintf("%.1f", 100 * abs(interaction_treatment_low_phq$estimate[1]))` percentage points** than waitlist (95% CI **`r sprintf("%.1f", 100 * interaction_treatment_low_phq$conf.low[1])` to `r sprintf("%.1f", 100 * interaction_treatment_low_phq$conf.high[1])` percentage points**). Within the baseline PHQ-9 `>= 10` category, the corresponding probability is **`r if (interaction_treatment_high_phq$estimate[1] >= 0) "higher" else "lower"` by `r sprintf("%.1f", 100 * abs(interaction_treatment_high_phq$estimate[1]))` percentage points** (95% CI **`r sprintf("%.1f", 100 * interaction_treatment_high_phq$conf.low[1])` to `r sprintf("%.1f", 100 * interaction_treatment_high_phq$conf.high[1])` percentage points**).

The associated therapist-guided-versus-waitlist odds ratios are **`r sprintf("%.2f", interaction_odds_ratio_low_phq$odds_ratio[1])`** (95% CI **`r sprintf("%.2f", interaction_odds_ratio_low_phq$odds_ratio_low[1])` to `r sprintf("%.2f", interaction_odds_ratio_low_phq$odds_ratio_high[1])`**) for PHQ-9 `< 10` and **`r sprintf("%.2f", interaction_odds_ratio_high_phq$odds_ratio[1])`** (95% CI **`r sprintf("%.2f", interaction_odds_ratio_high_phq$odds_ratio_low[1])` to `r sprintf("%.2f", interaction_odds_ratio_high_phq$odds_ratio_high[1])`**) for PHQ-9 `>= 10`. They compare odds, not probabilities. Use the conditional probability differences above—not these odds ratios—to describe the clinical contrast.
:::

```{r}
#| label: plot-interaction-probabilities
#| fig-cap: "Interaction-model estimated probabilities of treatment response. Error bars are 95% confidence intervals."
interaction_probabilities |>
  ggplot(aes(x = trt, y = estimate, group = phq_cat, colour = phq_cat)) +
  geom_point(position = position_dodge(width = 0.1)) +
  geom_line(position = position_dodge(width = 0.1)) +
  geom_errorbar(
    aes(ymin = conf.low, ymax = conf.high),
    width = 0.1,
    position = position_dodge(width = 0.1)
  ) +
  labs(
    title = "Interaction logistic model",
    subtitle = "Treatment probabilities may differ by baseline PHQ-9 category",
    x = "Treatment group",
    y = "Estimated probability of treatment response (95% CI)",
    colour = "Baseline PHQ-9"
  ) +
  theme_minimal()
```

```{r}
#| label: plot-conditional-treatment-effects
#| fig-cap: "Conditional therapist-guided-minus-waitlist probability differences from the interaction model. Error bars are 95% confidence intervals."
plot_comparisons(
  mod_interaction,
  variables = list(trt = c("waitlist", "therapist-guided")),
  by = "phq_cat",
  type = "response"
) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  labs(
    title = "Conditional treatment effects on the probability scale",
    x = "Baseline PHQ-9 category",
    y = "Therapist-guided minus waitlist probability difference (95% CI)"
  ) +
  theme_minimal()
```

# Interpreting the interaction

An interaction model can describe different conditional estimates, but it does not automatically establish a clinically meaningful subgroup. Before presenting a subgroup finding, ask:

1.  Was the moderator and interaction planned before looking at the data?
2.  Are the conditional probability differences precise enough to distinguish meaningful patterns from sampling variation?
3.  Does the interaction plot show a consistent, clinically interpretable pattern rather than an isolated coefficient or p-value?
4.  Would the conclusion be different if PHQ-9 were modelled continuously? See [logistic regression with a binary and numeric predictor](logistic-binary-numeric-predictors.qmd).

The interaction coefficient is a difference of **log-odds contrasts**. It is not itself the difference between the two probability differences. `avg_comparisons()` gives the quantities we want to communicate: therapist-guided-minus-waitlist probability differences within the explicitly stated PHQ-9 categories.

# Summary

In this chapter you learned to:

- derive `phq_cat` explicitly as baseline PHQ-9 `< 10` versus `>= 10`;
- fit additive and `trt * phq_cat` binary logistic models with `stats::glm(..., family = binomial)`;
- report treatment contrasts as conditional probability differences in percentage points with 95% CIs;
- distinguish a conditional odds ratio from a probability difference;
- distinguish effect modification, causal interaction, and moderation, and name the scale on which effect modification is assessed; and
- use predicted-probability and conditional-effect plots to inspect a proposed subgroup pattern.
