---
title: "Logistic regression with a numeric predictor"
toc: true
number-sections: true
---

# Numeric predictors and probabilities

A numeric predictor can change the probability of an outcome continuously. Here we model the probability of treatment response as a function of baseline GAD-7 score (`gad_screen`). The logistic model is linear on the log-odds scale, but its probability curve is generally curved.

```{r}
#| label: setup
#| message: false
#| warning: false
library(tidyverse)
library(here)
library(marginaleffects)

df_clean <- read_csv2(here("..", "..", "data", "steps_clean.csv")) |>
  filter(!is.na(lsas_post), !is.na(gad_screen)) |>
  mutate(
    lsas_response = factor(
      if_else(
        (lsas_screen - lsas_post) / lsas_screen >= 0.30,
        "Response",
        "No response"
      ),
      levels = c("No response", "Response")
    )
  )
```

```{r}
#| label: fit-numeric-model
mod_gad <- glm(
  lsas_response ~ gad_screen,
  family = binomial,
  data = df_clean
)
summary(mod_gad)

# Store the coefficients for the calculations below
b0 <- coef(mod_gad)[["(Intercept)"]]
b_gad <- coef(mod_gad)[["gad_screen"]]
```

The model coefficients tell us that the estimated log-odds when GAD-7 is 0 (Intercept) is `r sprintf("%.2f", b0)`, and that the estimated log-odds changes by `r sprintf("%.3f", b_gad)` for each one-point increase in GAD-7.

If we exponentiate the gad_screen estimate, we get the odds ratio of treatment response for each one-unit increase in GAD-7 at screening.

```{r}
#| label: numeric-model-manual-or
exp(b_gad)
```

We can also get the estimated odds ratio for an x-point increase in GAD-7 at screening, by using the same procedure while multiplying the coefficient by x. For example, if we wanted the odds ratio for a 10-point increase in GAD-7, we could do:

```{r}
#| label: numeric-model-manual-or-10-point
exp(b_gad * 10) # OR for response for a 10-point increase in GAD-7
```

Using the `avg_slopes()` function from the `marginaleffects` package, we can get these estimates on the probability scale, instead of the log-odds scale.

```{r}
#| label: numeric-model-probability
gad_slope <- avg_slopes(mod_gad, variables = "gad_screen", type = "response")
gad_slope
```

::: {.callout-tip collapse="true"}
## Model interpretation

A one-point higher baseline GAD-7 score changes the estimated probability of treatment response by **`r sprintf("%.2f", 100 * gad_slope$estimate[1])` percentage points on average** (95% CI **`r sprintf("%.2f", 100 * gad_slope$conf.low[1])`** to **`r sprintf("%.2f", 100 * gad_slope$conf.high[1])` percentage points**).

Because the logit link is nonlinear, a one-point probability change can differ across GAD-7 values. The coefficient in `summary()` is a log-odds change; exponentiating it gives an odds ratio, not this probability-scale slope.
:::

```{r}
#| label: plot-numeric-predictions
plot_predictions(mod_gad, condition = "gad_screen", type = "response") +
  labs(
    title = "Estimated probability of treatment response",
    subtitle = "Across baseline GAD-7 scores; 95% confidence interval",
    x = "GAD-7 score at screening",
    y = "Probability of treatment response"
  ) +
  theme_minimal()
```

# From a continuous score to a curve

A linear predictor does not imply a straight line on the probability scale. For this reason, plots and predictions at clinically meaningful values are usually more informative than reporting a coefficient alone.

```{r}
#| label: numeric-predictions-at-quartiles
gad_values <- quantile(
  df_clean$gad_screen,
  probs = c(.25, .50, .75),
  na.rm = TRUE
)

predictions(
  mod_gad,
  newdata = datagrid(gad_screen = gad_values),
  type = "response"
)
```

# Summary

- Logistic regression can model a binary outcome using a continuous predictor.
- Probability curves are usually easier to communicate than log-odds coefficients.
- An average probability slope is an interpretable summary, but it can conceal variation across the predictor range.
