---
title: "Multinomial logistic regression"
toc: true
number-sections: true
---

# Why a multinomial model?

The binary outcome used earlier distinguishes only response from non-response. Sometimes the question concerns how treatment redistributes probability across **three or more outcome categories**.

We use these teaching categories:

- `low_no_to_mild`: LSAS \< 50;
- `moderate_marked`: 50 ≤ LSAS \< 80; and
- `severe_very_severe`: LSAS ≥ 80.

They are a pedagogical severity grouping, not validated diagnostic or remission thresholds.

# Long data for `mlogit`

`mlogit` represents a multinomial outcome as a choice among alternatives. Each participant therefore appears once for every possible severity category. One and only one alternative is marked as the observed outcome.

```{r}
#| label: setup
#| message: false
#| warning: false
library(tidyverse)
library(here)
library(mlogit)
library(marginaleffects)

df_outcome <- read_csv2(here("..", "..", "data", "steps_clean.csv")) |>
  filter(!is.na(lsas_post), !is.na(trt)) |>
  mutate(
    severity_observed = case_when(
      lsas_post < 50 ~ "low_no_to_mild",
      lsas_post < 80 ~ "moderate_marked",
      TRUE ~ "severe_very_severe"
    ),
    severity_observed = factor(
      severity_observed,
      levels = c("low_no_to_mild", "moderate_marked", "severe_very_severe")
    ),
    trt = factor(trt,
      levels = c("waitlist", "therapist-guided", "self-guided")
    )
  ) |>
  select(id, trt, severity_observed)

long <- crossing(
  df_outcome,
  severity_alt = levels(df_outcome$severity_observed)
) |>
  mutate(choice = severity_alt == severity_observed)

long |>
  count(id) |>
  count(n)
```

```{r}
#| label: make-mlogit-data
mlogit_data <- mlogit.data(
  long,
  choice = "choice",
  shape = "long",
  chid.var = "id",
  alt.var = "severity_alt",
  drop.index = TRUE
)
```

# Fit a treatment model

```{r}
#| label: fit-multinomial-model
mod_multinomial <- mlogit(
  choice ~ 1 | trt,
  data = mlogit_data,
  reflevel = "low_no_to_mild"
)
summary(mod_multinomial)
```

## Interpreting the coefficients

As in binary logistic regression, the coefficients are on the **log-odds scale**. With three outcome categories, the model estimates two sets of coefficients because `low_no_to_mild` is the reference category:

- `moderate_marked` versus `low_no_to_mild`

- `severe_very_severe` versus `low_no_to_mild`

For example, the coefficient for therapist-guided treatment and `moderate_marked` is approximately -1.42. Exponentiating gives:

```{r}
#| label: multinomial-or
exp(-1.42)
```

This gives an odds ratio of approximately 0.24. In other words, the ratio between moderate/marked and low/no-to-mild outcomes is smaller in the therapist-guided group than in the waitlist group. More precisely, the odds of moderate/marked versus low/no-to-mild are 0.24 times as high in the therapist-guided group as in the waitlist group.

These odds ratios can become difficult to interpret because each compares **one outcome category with the reference outcome category**. We will therefore focus primarily on **model-based probabilities**, which show more directly how treatment redistributes the probability across the three severity categories.

```{r}
#| label: multinomial-probabilities
# Supply the ordinary long data frame, rather than mlogit_data, to
# marginaleffects when requesting predictions.
severity_probabilities <- avg_predictions(
  mod_multinomial,
  newdata = as.data.frame(long),
  by = "trt"
)
severity_probabilities
```

## Probability redistribution

The following custom hypothesis calculates therapist-guided minus waitlist probability differences separately for each severity category.

```{r}
#| label: multinomial-probability-differences
therapist_minus_waitlist <- function(x) {
  x |>
    group_by(.data$group, .data$trt) |>
    summarize(estimate = mean(.data$estimate), .groups = "drop") |>
    pivot_wider(names_from = "trt", values_from = "estimate") |>
    transmute(
      term = .data$group,
      estimate = .data[["therapist-guided"]] - .data$waitlist
    )
}

severity_differences <- predictions(
  mod_multinomial,
  newdata = as.data.frame(long),
  hypothesis = therapist_minus_waitlist
)
severity_differences
```

::: {.callout-tip collapse="true"}
## Model interpretation

Compared with waitlist, therapist-guided treatment changes the estimated probability of low/no-to-mild severity by **`r sprintf("%.1f", 100 * severity_differences$estimate[severity_differences$term == "low_no_to_mild"])` percentage points** (95% CI **`r sprintf("%.1f", 100 * severity_differences$conf.low[severity_differences$term == "low_no_to_mild"])`** to **`r sprintf("%.1f", 100 * severity_differences$conf.high[severity_differences$term == "low_no_to_mild"])` percentage points**).

The probability of severe/very severe symptoms changes by **`r sprintf("%.1f", 100 * severity_differences$estimate[severity_differences$term == "severe_very_severe"])` percentage points** (95% CI **`r sprintf("%.1f", 100 * severity_differences$conf.low[severity_differences$term == "severe_very_severe"])`** to **`r sprintf("%.1f", 100 * severity_differences$conf.high[severity_differences$term == "severe_very_severe"])` percentage points**). The category probabilities must sum to one, so an increase in one category is accompanied by decreases elsewhere.
:::

```{r}
#| label: plot-multinomial-probabilities
severity_probabilities |>
  ggplot(aes(x = trt, y = estimate, color = group, group = group)) +
  geom_point(position = position_dodge(width = .2)) +
  geom_errorbar(
    aes(ymin = conf.low, ymax = conf.high),
    width = .1,
    position = position_dodge(width = .2)
  ) +
  labs(
    title = "Estimated post-treatment severity probabilities",
    subtitle = "Multinomial logit; 95% confidence intervals",
    x = "Treatment group",
    y = "Estimated probability",
    color = "Severity category"
  ) +
  theme_minimal()
```

# Summary

- Multinomial logit models probabilities for multiple outcome categories.
- `mlogit` needs a long participant-by-alternative data structure.
- Interpret the model as a redistribution of outcome probabilities, not as isolated coefficient changes.
- The probability differences for the categories sum to zero across any treatment contrast.
