---
title: "Multinomial logistic regression"
toc: true
number-sections: true
---

# Why a multinomial model?

The binary outcome used earlier distinguishes only response from non-response. Sometimes the question concerns how treatment redistributes probability across **three or more outcome categories**.

We use these teaching categories:

- `low_no_to_mild`: LSAS \< 50;
- `moderate_marked`: 50 ≤ LSAS \< 80; and
- `severe_very_severe`: LSAS ≥ 80.

They are a pedagogical severity grouping, not validated diagnostic or remission thresholds.

::: callout-note
## Ordered categories

These categories are ordered: low/no-to-mild < moderate/marked < severe/very severe.
The multinomial model ignores that order.
It treats the categories as unordered labels and estimates a separate set of treatment coefficients for each non-reference category, so with two active treatment arms it has four treatment coefficients.
For an ordinal outcome, the usual first choice is a proportional-odds (cumulative logit) model, which uses the ordering and estimates one set of treatment coefficients for all categories, here two [@Agresti2010Ordinal].
It achieves this by assuming that the treatment odds ratio is the same at every cut-point between categories (low versus moderate or severe, and low or moderate versus severe), an assumption worth checking.
The cumulative model can be motivated by an underlying (latent) continuous severity that is cut at thresholds; here that is literally how the categories were made, from LSAS cut at 50 and 80.
When the full LSAS score is available, as it is here, analyzing it directly usually beats any categorical model, for the reasons given in the [foundations chapter](logistic-intro.qmd).

We use the multinomial model in this course because it is the general model for categorical outcomes, applies whether or not the categories are ordered, and makes the long-format data structure explicit.
The probability-scale interpretation you learn here, how treatment redistributes probability across the categories, carries over directly to ordinal models.
:::

# Long data for `mlogit`

`mlogit` represents a multinomial outcome as a choice among alternatives. Each participant therefore appears once for every possible severity category. One and only one alternative is marked as the observed outcome.

```{r}
#| label: setup
#| message: false
#| warning: false
library(tidyverse)
library(here)
library(mlogit)
library(marginaleffects)

df_outcome <- read_csv2(here("data", "steps_clean.csv")) |>
  filter(!is.na(lsas_post), !is.na(trt)) |>
  mutate(
    severity_observed = case_when(
      lsas_post < 50 ~ "low_no_to_mild",
      lsas_post < 80 ~ "moderate_marked",
      .default = "severe_very_severe"
    ),
    severity_observed = factor(
      severity_observed,
      levels = c("low_no_to_mild", "moderate_marked", "severe_very_severe")
    ),
    trt = factor(trt, levels = c("waitlist", "therapist-guided", "self-guided"))
  ) |>
  select(id, trt, severity_observed)

long <- crossing(
  df_outcome,
  severity_alt = levels(df_outcome$severity_observed)
) |>
  mutate(choice = severity_alt == severity_observed)

long |>
  count(id) |>
  count(n)
```

```{r}
#| label: make-mlogit-data
mlogit_data <- mlogit.data(
  long,
  choice = "choice",
  shape = "long",
  chid.var = "id",
  alt.var = "severity_alt",
  drop.index = TRUE
)
```

# Fit a treatment model

```{r}
#| label: fit-multinomial-model
mod_multinomial <- mlogit(
  choice ~ 1 | trt,
  data = mlogit_data,
  reflevel = "low_no_to_mild"
)
summary(mod_multinomial)

# Store one coefficient for the calculation below
b_multinomial <- coef(mod_multinomial)
b_guided_moderate <- b_multinomial[["trttherapist-guided:moderate_marked"]]
```

## Interpreting the coefficients

As in binary logistic regression, the coefficients are on the **log-odds scale**. With three outcome categories, the model estimates two sets of coefficients because `low_no_to_mild` is the reference category:

- `moderate_marked` versus `low_no_to_mild`

- `severe_very_severe` versus `low_no_to_mild`

For example, the coefficient for therapist-guided treatment and `moderate_marked` is approximately `r sprintf("%.2f", b_guided_moderate)`. Exponentiating gives:

```{r}
#| label: multinomial-or
exp(b_guided_moderate)
```

This gives an odds ratio of approximately `r sprintf("%.2f", exp(b_guided_moderate))`. In other words, the ratio between moderate/marked and low/no-to-mild outcomes is smaller in the therapist-guided group than in the waitlist group. More precisely, the odds of moderate/marked versus low/no-to-mild are `r sprintf("%.2f", exp(b_guided_moderate))` times as high in the therapist-guided group as in the waitlist group.

These odds ratios can become difficult to interpret because each compares **one outcome category with the reference outcome category**. We will therefore focus primarily on **model-based probabilities**, which show more directly how treatment redistributes the probability across the three severity categories.

```{r}
#| label: multinomial-probabilities
# marginaleffects cannot recover the training data from an mlogit model,
# so attach the ordinary long data frame explicitly. Supply the same data
# frame, rather than mlogit_data, when requesting predictions.
mod_multinomial <- set_modeldata(mod_multinomial, as.data.frame(long))

severity_probabilities <- avg_predictions(
  mod_multinomial,
  newdata = as.data.frame(long),
  by = "trt"
)
severity_probabilities
```

## Probability redistribution

The following custom hypothesis calculates therapist-guided minus waitlist probability differences separately for each severity category.

```{r}
#| label: multinomial-probability-differences
therapist_minus_waitlist <- function(x) {
  x |>
    group_by(.data$group, .data$trt) |>
    summarize(estimate = mean(.data$estimate), .groups = "drop") |>
    pivot_wider(names_from = "trt", values_from = "estimate") |>
    transmute(
      term = .data$group,
      estimate = .data[["therapist-guided"]] - .data$waitlist
    )
}

severity_differences <- predictions(
  mod_multinomial,
  newdata = as.data.frame(long),
  hypothesis = therapist_minus_waitlist
)
severity_differences
```

::: {.callout-tip collapse="true"}
## Model interpretation

Compared with waitlist, therapist-guided treatment changes the estimated probability of low/no-to-mild severity by **`r sprintf("%.1f", 100 * severity_differences$estimate[severity_differences$term == "low_no_to_mild"])` percentage points** (95% CI **`r sprintf("%.1f", 100 * severity_differences$conf.low[severity_differences$term == "low_no_to_mild"])`** to **`r sprintf("%.1f", 100 * severity_differences$conf.high[severity_differences$term == "low_no_to_mild"])` percentage points**).

The probability of severe/very severe symptoms changes by **`r sprintf("%.1f", 100 * severity_differences$estimate[severity_differences$term == "severe_very_severe"])` percentage points** (95% CI **`r sprintf("%.1f", 100 * severity_differences$conf.low[severity_differences$term == "severe_very_severe"])`** to **`r sprintf("%.1f", 100 * severity_differences$conf.high[severity_differences$term == "severe_very_severe"])` percentage points**). The category probabilities must sum to one, so an increase in one category is accompanied by decreases elsewhere.
:::

```{r}
#| label: plot-multinomial-probabilities
severity_probabilities |>
  ggplot(aes(x = trt, y = estimate, color = group, group = group)) +
  geom_point(position = position_dodge(width = .2)) +
  geom_errorbar(
    aes(ymin = conf.low, ymax = conf.high),
    width = .1,
    position = position_dodge(width = .2)
  ) +
  labs(
    title = "Estimated post-treatment severity probabilities",
    subtitle = "Multinomial logit; 95% confidence intervals",
    x = "Treatment group",
    y = "Estimated probability",
    color = "Severity category"
  ) +
  theme_minimal()
```

# Summary

- Multinomial logit models probabilities for multiple outcome categories.
- `mlogit` needs a long participant-by-alternative data structure.
- Interpret the model as a redistribution of outcome probabilities, not as isolated coefficient changes.
- The probability differences for the categories sum to zero across any treatment contrast.
- Multinomial logit ignores any ordering of the categories; for an ordinal outcome, a proportional-odds model is usually the first choice.
