---
title: "Probability rules"
---

In this lab, we will look at how we can work with probability rules in R.

# Load packages and data

```{r}
#| label: load-packages
#| message: false

library(tidyverse)
library(here)
library(knitr)

d_bl <- read_rds(here("data", "steps_baseline.rds"))
```

# Probability rules

Let's look at how we can implement probability rules in R and apply them to our dataset. In the [Import and clean data](import-clean.qmd) lab, we created some categorical variables at baseline. These were just simulated, so we will not try to make sense of the results.

```{r}
#| label: glimpse-data

d_bl |>
  select(where(is.character)) |>
  glimpse()
```

For this lab, we will use the "education" and "income" variables. First, we will create summaries of these variables.

```{r}
#| label: tbl-edu-summary
#| tbl-cap: "Education levels"

edu_summary <- d_bl |>
  group_by(education) |>
  summarise(
    n = n(),
    proportion = n / nrow(d_bl),
    percent = round(proportion * 100, 1)
  )

kable(edu_summary)
```

```{r}
#| label: tbl-income-summary
#| tbl-cap: "Income levels"

income_summary <- d_bl |>
  group_by(income) |>
  summarise(
    n = n(),
    proportion = n / nrow(d_bl),
    percent = round(proportion * 100, 1)
  )

kable(income_summary)
```

## Check that the sum of all probabilities is 1

We do a quick check of the *education* variable, which has three levels: "Primary", "Secondary", and "University". When we count the proportion of each level, we get the following: 

The proportion with the "Primary" level is `r edu_summary |> filter(education == "Primary") |> pull(proportion)`, the proportion with the "Secondary" level is `r edu_summary |> filter(education == "Secondary") |> pull(proportion)`, and the proportion with the "University" level is `r edu_summary |> filter(education == "University") |> pull(proportion)`. These add up to `r sum(edu_summary$proportion)`. All good!

## Complement rule

The probability of an event not occurring is 1 minus the probability that it will occur.

Let's check this for the "Secondary" level.

$$
P(\text{not Secondary}) = 1 - P(\text{Secondary})
$$

```{r}
#| label: complement-rule

# probability of Secondary education
p_secondary <- edu_summary |>
  filter(education == "Secondary") |>
  pull(proportion)

# complement rule
p_not_secondary <- 1 - p_secondary
```

- P(Secondary education) = `r round(p_secondary, 4)`
- P(not Secondary education) = `r round(p_not_secondary, 4)`
- Check complement rule, sum = `r round(p_secondary + p_not_secondary, 4)`


## Addition rule

The probability that event A or event B occurs (or both).

$$P(A \cup B) = P(A) + P(B) - P(A \cap B)$$

Let's implement this using our dataset. We'll look at the probability of having either "Primary" education OR being in the "Low" income group.

```{r}
#| label: addition-rule

# get probabilities
p_primary <- edu_summary |>
  filter(education == "Primary") |>
  pull(proportion)

p_low_income <- income_summary |>
  filter(income == "Low") |>
  pull(proportion)

# calculate probability of both Primary education AND Low income
p_both <- d_bl |>
  filter(education == "Primary" & income == "Low") |>
  nrow() / nrow(d_bl)

# addition rule
p_either <- p_primary + p_low_income - p_both
```

- P(Primary education) = `r round(p_primary, 4)`
- P(Low income) = `r round(p_low_income, 4)`
- P(Both) = `r round(p_both, 4)`
- P(Either) = `r round(p_either, 4)`

## Multiplication rule

For independent events, the probability of both events occurring is the product of their individual probabilities:

$$P(A \cap B) = P(A) \times P(B)$$

For dependent events, we need to account for the conditional probability:

$$P(A \cap B) = P(A) \times P(B|A)$$

Let's check if education and income are independent by comparing the observed joint probability with the product of marginal probabilities. We will first use `group_by()` and `summarise()` to create @tbl-edu-income.

```{r}
#| label: tbl-edu-income
#| tbl-cap: "Joint probabilities of education and income"

edu_income_table <- d_bl |>
  group_by(education, income) |>
  summarise(
    n = n(),
    proportion = n / nrow(d_bl),
    percent = round(proportion * 100, 1),
    .groups = "drop"
  )

kable(edu_income_table)
```

```{r}
#| label: independence-check

# Check independence for Primary education and Low income
p_primary_indep <- p_primary * p_low_income
p_primary_dep <- p_both
```

- If independent: P(Primary ∩ Low income) = `r round(p_primary_indep, 6)`
- Observed: P(Primary ∩ Low income) = `r round(p_primary_dep, 6)`
- Difference = `r round(abs(p_primary_indep - p_primary_dep), 6)`

::: {.callout-tip}
Keep in mind that this is simulated data, so the numbers may not represent the real world. Nonetheless, if we observed a result like this, what would we conclude?
:::

## Conditional probability

The probability of event B occurring given that event A has occurred:

$$P(B|A) = \frac{P(A \cap B)}{P(A)}$$

Let's calculate the probability of having "Low" income given that someone has "Primary" education:

```{r}
#| label: conditional-probability

p_low_given_primary <- p_both / p_primary

```

- P(Low income | Primary education) = `r round(p_low_given_primary, 4)`
