---
title: "Logistic regression with binary and categorical predictors"
format:
  html:
    df-print: kable
toc: true
execute:
  eval: false
---

# Scenario

Your supervisor has drafted an abstract claiming that one STePS treatment arm is clearly better than another. The only result in the draft is a logistic-regression coefficient.

Prepare the analysis for a meeting later today. Decide which treatment contrast is clinically useful, state why, and make the uncertainty visible before anyone writes “paradigm-shifting.”

# Your deliverable

Create a group Quarto notebook with these sections:

1. **Question and estimand** — Choose one clinically relevant contrast among
   waitlist, therapist-guided treatment, and self-guided treatment. State the
   direction explicitly (for example, therapist-guided minus waitlist) and say
   why that comparison is useful.
2. **Outcome and data check** — Define `lsas_post_bin` as `< 50` versus `>= 50`.
   State that this is a teaching/analysis threshold rather than a diagnostic or
   remission definition. Report the number of participants and outcome counts
   in each treatment arm, including how you handled missing post-treatment LSAS
   values.
3. **Model** — Fit a three-arm binary logistic model. Identify the reference
   category and the event whose probability is modelled.
4. **Results** — Report model-derived probabilities for the relevant treatment
   arms and your chosen, directed probability difference in percentage points
   with a 95% CI. If you include an odds ratio, name its comparison and explain
   that it is a ratio of odds, not a probability difference.
5. **Figure** — Include an uncertainty-aware probability-scale figure with 95%
   CIs. A plot of all three arm-specific probabilities is usually informative;
   add a contrast plot if it makes your chosen comparison clearer.
6. **Conclusion** — Give a short, plain-language clinical interpretation of the
   numerical probability difference and its uncertainty. State what the model
   alone does not establish.

# Starter code

Use this only as a starting point. Your group decides which contrast answers
your question, how to display it, and which limitations deserve emphasis.

```{r}
#| label: setup
#| message: false
#| warning: false
library(tidyverse)
library(here)
library(broom)
library(marginaleffects)

options(
  marginaleffects_print_omit = c("s.value"),
  marginaleffects_print_style = ifelse(interactive(), "summary", "tinytable")
)

df_categorical <- read_csv2(here("data", "steps_clean.csv")) |>
  filter(!is.na(lsas_post)) |>
  mutate(
    # The second level is the event modelled by binomial glm().
    lsas_post_bin = factor(
      if_else(lsas_post < 50, "Low", "High"),
      levels = c("High", "Low")
    ),
    trt = factor(
      trt,
      levels = c("waitlist", "therapist-guided", "self-guided")
    )
  )
```

```{r}
#| label: check-treatment-outcomes
# Decide what missing outcome values mean for the analysis before fitting a model.
df_categorical |>
  count(trt, lsas_post_bin, .drop = FALSE)

contrasts(df_categorical$trt)
```

```{r}
#| label: three-arm-logistic-model
#| message: false
#| warning: false
mod_categorical_lab <- stats::glm(
  lsas_post_bin ~ trt,
  family = binomial,
  data = df_categorical
)

# Replace these arms with the directed clinical contrast your group selected.
chosen_arms <- c("waitlist", "therapist-guided")

p_categorical_lab <- avg_predictions(
  mod_categorical_lab,
  by = "trt",
  type = "response"
)

chosen_difference <- avg_comparisons(
  mod_categorical_lab,
  variables = list(trt = chosen_arms),
  type = "response"
)

default_or <- tidy(
  mod_categorical_lab,
  conf.int = TRUE,
  exponentiate = TRUE
) |>
  filter(term == "trttherapist-guided")
```

::: {.callout-tip collapse="true"}
## Model interpretation template

Replace this paragraph with your own numerical interpretation. State the two
estimated probabilities, the directed probability difference in **percentage
points**, and its 95% CI. Then state the comparison and any conditioning
choices clearly.

If you report an odds ratio, give its 95% CI and explain that it compares odds,
not probabilities. Do not describe it as a probability ratio or as the
percentage-point difference.
:::

```{r}
#| label: probability-scale-figure
plot_predictions(mod_categorical_lab, by = "trt", type = "response") +
  labs(
    title = "Estimated probability of low post-treatment LSAS",
    subtitle = "Three-arm logistic regression; 95% confidence intervals",
    x = "Treatment group",
    y = "Probability of low post-treatment LSAS"
  ) +
  theme_minimal()
```

# Questions to discuss

- Which treatment contrast would change a real clinical or service decision?
  Which contrast merely makes a more dramatic abstract title?
- Does a reference-category coefficient answer the same question as your chosen
  probability difference? Why or why not?
- What can an odds ratio tell a reader, and why can it not be read as a
  percentage-point change?
- The model is unadjusted for baseline severity. Is that appropriate for your
  chosen estimand, and what would change if you adjusted for a pre-treatment
  variable?
- What information is lost when the continuous LSAS score is dichotomised?

# Before presenting

- [ ] Our contrast names the first group, second group, and direction.
- [ ] We define the modelled event and do not call `< 50` a diagnostic or
      remission threshold.
- [ ] We report group probabilities and a percentage-point probability
difference with a 95% CI.
- [ ] Our figure has a probability-scale y-axis and makes its 95% CIs clear.
- [ ] If we report an odds ratio, we identify it as a ratio of odds rather than
a probability difference.
- [ ] Our conclusion is clinically intelligible without relying on a log-odds
coefficient.
