---
title: "Lab 5: The continuous moderator that wanted to be a headline"
format:
  html:
    df-print: kable
toc: true
---

# Scenario

A senior collaborator has noticed that the interaction-model curve appears to
move across baseline PHQ-9 scores. Their draft conference abstract now contains
the sentence:

> *Therapist guidance is especially effective for the patients who need it most,
> as demonstrated by a continuous PHQ-9 moderation effect.*

The collaborator would like the figure “cleaned up” before the abstract is sent.
Your group is asked to evaluate the claim—not merely to make a graph that agrees
with it. The request is intentionally under-specified, as befits a result that
has acquired a life of its own during a PhD meeting.

# Your task

Prepare a short, reproducible Quarto notebook that evaluates whether the claim
is supported by a binary logistic model of the probability of a low
post-treatment LSAS score. Work with the waitlist and therapist-guided treatment
arms. Treat `phq9_screen` as continuous; do not turn it into a median split just
to make the narrative more convenient.

You may choose reasonable presentation details, but your notebook must make each
analysis decision visible and must distinguish what the model estimates from what
the collaborator would like it to establish.

# Required deliverables

Your notebook must contain all of the following.

1. **Question and estimands**
   - State the binary outcome and the event whose probability is modelled.
   - Define the direction of the treatment contrast as therapist-guided minus
     waitlist.
   - State one conditional estimand at a meaningful PHQ-9 value and one
     population-average estimand over the observed PHQ-9 distribution.
2. **Data and model choices**
   - Define `lsas_post_bin` as `< 50` versus `>= 50`; name this as a teaching
     analysis threshold rather than a diagnostic or remission definition.
   - Set waitlist as the reference group, retain only the two treatment arms,
     and report how missing outcome or PHQ-9 values were handled.
   - Fit both an additive logistic model and an interaction logistic model using
     `stats::glm(..., family = binomial)`.
3. **Probability-scale results**
   - Report therapist-guided-minus-waitlist probability differences with 95% CIs
     at at least three prespecified PHQ-9 values within the observed range.
     Explain why those values are meaningful for your analysis; do not select
     them because they make the curve look most dramatic.
   - Estimate the interaction model's population-average treatment probability
     difference, with a 95% CI, by averaging over the observed PHQ-9 values.
   - Include an uncertainty-aware curve of the conditional probability
     difference across PHQ-9. A plot of group-specific probabilities is useful
     too, but does not replace the direct contrast plot.
4. **Interpretation and conclusion**
   - Write a short, clinically intelligible conclusion that gives numerical
     percentage-point differences, their intervals, their conditioning values,
     and the direction of the contrast.
   - Explain whether the data support the proposed moderator claim. Do not use a
     visually sloping point-estimate curve as evidence by itself.
   - If you report an odds ratio, identify the PHQ-9 value to which it applies
     and explain why it is a ratio of odds, not a probability difference.

# Starter code

This setup is deliberately not a complete analysis. It makes the outcome and
reference groups explicit; your group must decide which contrasts, tables,
figures, diagnostics, and wording answer the question.

```{r}
#| label: setup
#| message: false
#| warning: false
library(tidyverse)
library(here)
library(marginaleffects)

df_binary <- read_csv2(here("data", "steps_clean.csv")) |>
  mutate(
    # The second factor level is the event modelled by binomial glm().
    lsas_post_bin = factor(
      if_else(lsas_post < 50, "Low", "High"),
      levels = c("High", "Low")
    ),
    trt = factor(
      trt,
      levels = c("waitlist", "therapist-guided", "self-guided")
    )
  ) |>
  filter(trt %in% c("waitlist", "therapist-guided")) |>
  drop_na(lsas_post_bin, phq9_screen) |>
  droplevels()
```

```{r}
#| label: models
mod_additive <- stats::glm(
  lsas_post_bin ~ trt + phq9_screen,
  family = binomial,
  data = df_binary
)

mod_interaction <- stats::glm(
  lsas_post_bin ~ trt * phq9_screen,
  family = binomial,
  data = df_binary
)

# Which probability-scale contrasts answer your stated estimands?
```

# Questions for the group

- What is the difference between a treatment contrast at PHQ-9 = 10 and a
  treatment contrast averaged over the observed PHQ-9 distribution?
- In an additive logistic model, does a constant treatment log-odds difference
  require a constant treatment probability difference? Use predictions or
  comparisons to justify your answer.
- How much support for effect modification is visible once the conditional
  probability-difference curve includes its 95% confidence interval?
- What population would your population-average contrast describe? Would it be
  unchanged in a clinic with a different baseline PHQ-9 distribution?
- Which parts of the collaborator's sentence are empirical findings, and which
  are causal or clinical claims that this analysis alone cannot establish?

# Before presenting

- [ ] The Quarto notebook runs from top to bottom in a clean R session.
- [ ] The analysis states the modelled event, treatment reference group, and
      direction of every treatment contrast.
- [ ] Both fitted models use `stats::glm(..., family = binomial)`;
      probability-scale results use `marginaleffects` rather than interpreting log-odds coefficients as the
      final clinical answer.
- [ ] The conditional contrasts, curve, and population-average contrast all
      include 95% confidence intervals.
- [ ] The conclusion reports numerical percentage-point results and does not
      turn an odds ratio into a probability difference.
- [ ] The conclusion distinguishes statistical uncertainty from a claim that a
      particular subgroup "needs" or benefits more from therapist guidance.
