---
title: "Logistic-regression foundations"
format:
  html:
    df-print: kable
toc: true
---

# Scenario

Your supervisor has received reviewer comments asking for a clinically intuitive binary outcome. Could the revised analysis show whether therapist guidance increases the chance of treatment response?

Prepare a short Quarto analysis for the next meeting. The request is intentionally incomplete, so decide what quantity answers it and make your assumptions visible.

# Your deliverable

Create a Quarto notebook with these sections:

1. **Question and outcome definition** — State the treatment comparison and define `lsas_response` as a reduction of at least 30% from baseline LSAS. Explain why this is a response outcome rather than a diagnostic or remission definition.
2. **Data check** — Report the number of participants and the outcome count in each treatment group. State how you handled missing post-treatment LSAS values.
3. **Model** — Fit a binary logistic model comparing therapist-guided treatment with waitlist. Clearly name the reference group and the event whose probability is modelled.
4. **Results** — Report estimated probabilities for both groups and the therapist-guided-minus-waitlist probability difference with a 95% CI. You may also report an odds ratio, but explain why it is not the same estimand as the probability difference.
5. **Figure** — Include one probability-scale figure with 95% CIs.
6. **Conclusion** — Write a short clinical interpretation. State the numerical percentage-point difference, its uncertainty, and what the model does *not* establish.

# Starter code

Use this only as a starting point. You are responsible for deciding what to show and how to explain it.

```{r}
#| label: setup
#| message: false
#| warning: false
library(tidyverse)
library(here)
library(marginaleffects)

df_binary <- read_csv2(here("data", "steps_clean.csv")) |>
  filter(!is.na(lsas_screen), !is.na(lsas_post)) |>
  mutate(
    lsas_response = factor(
      if_else((lsas_screen - lsas_post) / lsas_screen >= 0.30, "Response", "No response"),
      levels = c("No response", "Response")
    ),
    trt = factor(
      trt,
      levels = c("waitlist", "therapist-guided", "self-guided")
    )
  ) |>
  filter(trt %in% c("waitlist", "therapist-guided")) |>
  droplevels()
```

```{r}
#| label: model
mod_trt <- glm(
  lsas_response ~ trt,
  family = binomial,
  data = df_binary
)

# What probabilities and contrast answer your question?
```

# Questions to discuss

- What would change if `Low` were defined using a different threshold?
- Is the probability difference more useful for this clinical question than the odds ratio? Why?
- Does a treatment-group probability difference establish that therapist guidance caused the difference? What features of the trial design support or limit that interpretation?
- What information did the binary outcome discard from the original LSAS scale?

# Before presenting

- [ ] The notebook runs from top to bottom in a clean R session.
- [ ] The direction of the treatment contrast is unambiguous.
- [ ] The result is stated in percentage points with a 95% CI.
- [ ] The figure has a meaningful title and probability-scale y-axis.
- [ ] The conclusion does not describe an odds ratio as a probability difference.
