---
title: "Logistic-regression foundations"
format:
  html:
    df-print: kable
toc: true
---

# Scenario

Your supervisor has received reviewer comments asking for a clinically intuitive binary outcome.
Could the revised analysis show whether therapist guidance increases the chance of treatment response?

Prepare an analysis in a Quarto notebook for the next meeting.

# Your task

Create a Quarto notebook with these sections:

1. **Question and outcome definition** — State the treatment comparison and define `lsas_response` as a reduction of at least 30% from baseline LSAS.
2. **Data check** — Report the number of participants and the outcome count in each treatment group.
   State how you handled missing post-treatment LSAS values.
3. **Model** — Fit a binary logistic model comparing therapist-guided treatment with waitlist.
   Clearly name the reference group and the event whose probability is modeled.
4. **Results** — Report estimated probabilities for both groups and the therapist-guided-minus-waitlist probability difference with a 95% CI.
   Also report an odds ratio of the group comparison.
5. **Figure** — Include one probability-scale figure with 95% CIs.
6. **Conclusion** — Write a short clinical interpretation.
   State the numerical percentage-point difference, its uncertainty, and what the model does *not* establish.

# Starter code

Use this only as a starting point.
You are responsible for deciding what to show and how to explain it.

```{r}
#| label: setup
#| message: false
#| warning: false
library(tidyverse)
library(here)
library(marginaleffects)

df_binary <- read_csv2(here("data", "steps_clean.csv")) |>
  filter(!is.na(lsas_screen), !is.na(lsas_post)) |>
  mutate(
    lsas_response = factor(
      if_else(
        (lsas_screen - lsas_post) / lsas_screen >= 0.30,
        "Response",
        "No response"
      ),
      levels = c("No response", "Response")
    ),
    trt = factor(
      trt,
      levels = c("waitlist", "therapist-guided", "self-guided")
    )
  ) |>
  filter(trt %in% c("waitlist", "therapist-guided")) |>
  droplevels()
```

```{r}
#| label: model
mod_trt <- glm(
  lsas_response ~ trt,
  family = binomial,
  data = df_binary
)

# What probabilities and contrast answer your question?
```
